---
title: 'Forecasting Unemployment Trends: A Comparative Time Series Analysis of Colombia
  and the United States'
author: "Aye Nyein Thu, Mazhar Bhuyan, Yuqi Yang, Jisup Kwak"
date: "2025-03-24"
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---

## Setting R code chunk options

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  error = TRUE,   
  results = 'markup',
  tidy.opts = list(width.cutoff = 80),
  tidy = FALSE
)

```

## Loading packages and initializing
```{r packages, warning=FALSE, message=FALSE}
# Load required packages
library(readxl)
library(openxlsx)
library(writexl)
library(dplyr)
library(lubridate)
library(ggplot2)
library(cowplot)
library(forecast)  
library(Kendall)
library(tseries)
library(outliers)
library(tidyverse)
library(smooth)
library(trend)
library(kableExtra)
library(tidyr)
library(gt) #install.packages("gt")
library(gridExtra) #install.packages("gridExtra")
library(zoo)
library(imputeTS)


# Check working directory
getwd()
```

## Importing and Wrangling Data
```{r importing data, results='hide', warning=FALSE, message=FALSE}
## Raw Data Set: Unemployment Rate by Age (Thousands)
# Import data set
UEAge.Thou <- read_excel(
  path="./Data/Raw/UE_Age(Thousands).xlsx", sheet = "Sheet1", col_names = TRUE)

# Format data set
UEAge.Thou_Processed <- UEAge.Thou %>%
  mutate(
    Month = ym(sub("M", "-", Month)), 
    Age15to24.Thou = as.numeric(`15-24`), 
    Age25above.Thou = as.numeric(`25+`),   
    AgeTotal.Thou = as.numeric(`15+`)) %>% 
  rename(Country="Reference area") %>% 
  select(Country,Month,Age15to24.Thou, Age25above.Thou, AgeTotal.Thou) %>% 
  arrange(Country, Month)

## Raw Data Set: Unemployment Rate by Age (%)
# Import data set
UEAge.Per <- read_excel(
  path="./Data/Raw/UE_Age(%).xlsx", sheet = "Sheet1", col_names = TRUE)

# Format data set
UEAge.Per_Processed <- UEAge.Per %>%
  mutate(
    Month = ym(sub("M", "-", Month)), 
    Age15to24.Per = as.numeric(`15-24`), 
    Age25above.Per = as.numeric(`25+`),   
    AgeTotal.Per = as.numeric(`15+`)) %>% 
  rename(Country="Reference area") %>% 
  select(Country,Month,Age15to24.Per, Age25above.Per, AgeTotal.Per) %>% 
  arrange(Country, Month)

## Raw Data Set: Unemployment Rate by Gender (Thousands)
# Import data set
UEGender.Thou <- read_excel(
  path="./Data/Raw/UE_Gender(Thousands).xlsx", sheet = "Sheet1", col_names = TRUE)

# Format data set
UEGender.Thou_Processed <- UEGender.Thou %>%
  mutate(
    Month = ym(sub("M", "-", Month)), 
    Female.Thou = as.numeric(Female), 
    Male.Thou = as.numeric(Male),   
    Total.Thou = as.numeric(Total)) %>% 
  rename(Country="Reference area") %>% 
  select(Country,Month,Female.Thou, Male.Thou, Total.Thou) %>% 
  arrange(Country, Month)

## Raw Data Set: Unemployment Rate by Gender (%)
# Import data set
UEGender.Per <- read_excel(
  path="./Data/Raw/UE_Gender(%).xlsx", sheet = "Sheet1", col_names = TRUE)

# Format data set
UEGender.Per_Processed <- UEGender.Per %>%
  mutate(
    Month = ym(sub("M", "-", Month)), 
    Female.Per = as.numeric(Female), 
    Male.Per = as.numeric(Male),   
    Total.Per = as.numeric(Total)) %>% 
  rename(Country="Reference area") %>% 
  select(Country,Month,Female.Per, Male.Per, Total.Per) %>% 
  arrange(Country, Month)
```

```{r data wrangling Colombia, warning=FALSE, message=FALSE}
# Combine all processed data sets 
UE_Countries <- UEAge.Thou_Processed %>% 
  left_join(UEAge.Per_Processed, by=c("Country", "Month")) %>% 
  left_join(UEGender.Thou_Processed, by=c("Country", "Month")) %>% 
  left_join(UEGender.Per_Processed, by=c("Country", "Month")) 

# Extract Colombia Data
Colombia <- UE_Countries %>% 
  filter(Country == "Colombia") %>% 
  select(-Country, AgeTotal.Per, AgeTotal.Thou) %>% 
  select(Month, Age15to24.Per, Age25above.Per, Female.Per, Male.Per,
         Total.Per, Age15to24.Thou, Age25above.Thou, Female.Thou, Male.Thou,
         Total.Thou) 


# Check Missing Value 
sum(is.na(Colombia))

# Extract US Data 
US <- UE_Countries %>% 
  filter(Country == "United States of America",
         Month >= as.Date("2001-01-01") & Month <= as.Date("2024-12-01")) %>% 
  select(-Country, AgeTotal.Per, AgeTotal.Thou) %>% 
  select(Month, Age15to24.Per, Age25above.Per, Female.Per, Male.Per,
         Total.Per, Age15to24.Thou, Age25above.Thou, Female.Thou, Male.Thou,
         Total.Thou) 

# Check Missing Value 
sum(is.na(US))
```

```{r data wrangling global, warning=FALSE, message=FALSE}
# Generate global unemployment data using simple average 
UE_Global <- UE_Countries %>%
  group_by(Month) %>%
  summarise(
    Age15to24.Thou = mean(`Age15to24.Thou`, na.rm = TRUE),
    Age25above.Thou = mean(`Age25above.Thou`, na.rm = TRUE), 
    AgeTotal.Thou = mean(`AgeTotal.Thou`, na.rm = TRUE), 
    Age15to24.Per = mean(`Age15to24.Thou`, na.rm = TRUE),
    Age25above.Per = mean(`Age25above.Per`, na.rm = TRUE), 
    AgeTotal.Per = mean(`AgeTotal.Per`, na.rm = TRUE), 
    Female.Thou = mean(`Female.Thou`, na.rm = TRUE),
    Male.Thou = mean(`Male.Thou`, na.rm = TRUE),
    Total.Thou = mean(`Total.Thou`, na.rm = TRUE),
    Female.Per = mean(`Female.Per`, na.rm = TRUE),
    Male.Per = mean(`Male.Per`, na.rm = TRUE),
    Total.Per = mean(`Total.Per`, na.rm = TRUE))

# Generate global unemployment data using weighted average 
UE_Global_Weighted <- UE_Countries %>%
  filter(apply(UE_Countries[, 3:14], 1, function(x) all(!is.na(x)))) %>%  
  group_by(Month) %>%
  summarise(
    Age15to24.Per = sum(Age15to24.Thou) / sum(Age15to24.Thou / Age15to24.Per), 
    Age25above.Per = sum(Age25above.Thou) / sum(Age25above.Thou / Age25above.Per),
    Female.Per = sum(Female.Thou) / sum(Female.Thou / Female.Per),
    Male.Per = sum(Male.Thou) / sum(Male.Thou / Male.Per),
    Total.Per = sum(Total.Thou) / sum(Total.Thou / Total.Per),
    
    Age15to24.Thou = mean(Age15to24.Thou, na.rm = TRUE),
    Age25above.Thou = mean(Age25above.Thou, na.rm = TRUE),
    Female.Thou = mean(Female.Thou, na.rm = TRUE),
    Male.Thou = mean(Male.Thou, na.rm = TRUE),
    Total.Thou = mean(Total.Thou, na.rm = TRUE)
  )

# Generate weighted global unemployment yearly
UE_Global_Yearly <- UE_Global_Weighted %>% 
  mutate(Year = year(Month)) %>%
  group_by(Year) %>% 
  summarize(across(where(is.numeric), mean)) #%>% 
  #filter(Year >= 2000 & Year <= 2024)
```

## Descriptive Statistics 
### Global Unemployment Status
```{r global, warning=FALSE, message=FALSE}
# Check total no. of countries 
print(length(unique(UE_Countries$Country)))

# Plot global unemployment rate for simple average vs. weighted average 
UE_Global_Combined <- bind_rows(
  UE_Global %>% mutate(Source = "Simple Average"),
  UE_Global_Weighted %>% mutate(Source = "Weighted Average")
)

ggplot(UE_Global_Combined, aes(x = Month, y = Total.Per, color = Source)) +
  geom_line() +               
  labs(title = "Global Unemployment Rate: Simple Vs. Weighted Average",
       x = "Month", y = "Global Unemployment Rate (%)", color = "Source") 

# Plots
# Global Unemployment Rate 
ggplot(UE_Global_Yearly, aes(x = Year, y = Total.Per)) +
  geom_line() +               
  labs(title = "Global Unemployment Rate",
       x = "Year", y = "Global Unemployment Rate (%)")

# Global Unemployment Rate by Age
ggplot(UE_Global_Yearly, aes(x = Year)) +
  geom_line(aes(y = Age15to24.Per, color = "Age 15-24")) +
  geom_line(aes(y = Age25above.Per, color = "Age 25+")) +
  labs(title = "Global Unemployment Rate by Age",
       x = "Year", y = "Global Unemployment Rate (%)", color = "Age Group") +
  scale_color_manual(values = c("Age 15-24" = "blue", "Age 25+" = "red"))

# Global Unemployment Rate by Sex
ggplot(UE_Global_Yearly, aes(x = Year)) +
  geom_line(aes(y = Female.Per, color = "Female")) +
  geom_line(aes(y = Male.Per, color = "Male")) +
  labs(title = "Global Unemployment Rate by Sex",
       x = "Year", y = "Global Unemployment Rate (%)", color = "Sex Group") +
  scale_color_manual(values = c("Female" = "blue", "Male" = "red"))
```

### Unemployment Status in Colombia and United States
Starting in January 2022, Colombia revised its methodology for compiling labor statistics. As a result, caution is needed when analyzing employment data across this period.

```{r Colombia, warning=FALSE, message=FALSE}
## Unemployment Status in Colombia
p1_PE <- ggplot(Colombia, aes(x = Month, y = Total.Thou)) +
  geom_line(color = "blue", alpha = 1) +
  labs(x = "", y = "Unemployment (Thousands)") 

Colombia_LF <- Colombia %>% 
  mutate(LF.Part = (Total.Thou / Total.Per) * 100) 

p2_PE <- ggplot(Colombia_LF, aes(x = Month, y = LF.Part)) +
  geom_line(color = "orange", alpha = 1) +
  labs(x = "", y = "Laborforce Participation (Thousands)")

p3_PE <- ggplot(Colombia, aes(x = Month, y = Total.Per)) +
  geom_line(color = "red", alpha = 1) +
  labs(x = "Year", y = "Unemployment Rate (%)") 

grid.arrange(p1_PE, p2_PE, p3_PE, ncol = 3,
             top = "Unemployment Status in Colombia")


## US Unemployment Status in US 
p1_US <- ggplot(US, aes(x = Month, y = Total.Thou)) +
  geom_line(color = "blue", alpha = 1) +
  labs(x = "", y = "Unemployment (Thousands)") 

p2_US <- ggplot(US, aes(x = Month, y = Total.Per)) +
  geom_line(color = "red", alpha = 1) +
  labs(x = "Year", y = "Unemployment Rate (%)") 

US_LF <- US %>% 
  mutate(LF.Part = (Total.Thou / Total.Per) * 100) 

p3_US <- ggplot(US_LF, aes(x = Month, y = LF.Part)) +
  geom_line(color = "orange", alpha = 1) +
  labs(x = "", y = "Laborforce Participation (Thousands)")

# Arrange plots side by side
grid.arrange(p1_PE, p2_PE, p3_PE, ncol = 3,
             top = "Unemployment Status in the US")

# Comparison of Unemployment Rate between Colombia and US
UE_Comparison <- UE_Countries %>% 
  filter(Country %in% c("Colombia", "United States of America")) %>% 
  mutate(Month = as.Date(Month)) %>% 
  filter(Month >= as.Date("2001-01-01") & Month <= as.Date("2024-12-01"))

ggplot(UE_Comparison, aes(x = Month, y = Total.Per, color = Country)) +
  geom_line(size = 0.8) +  
  geom_point(size = 0.8, alpha = 0.7) + 
  theme_gray(base_size = 14) +  
  labs(title = "Unemployment Rate by Countries (2001-2024)",
    x = "Year", y = "Unemployment Rate (%)", color = "Country") +
  theme(plot.title = element_text(face = "bold", size = 16),
    axis.title.y = element_text(size = 14), axis.text = element_text(size = 12),
    legend.position = "right")

# Comparison of Unemployment Rate by Age Group 
PE1 <- ggplot(Colombia, aes(x = Month)) +
  geom_line(aes(y = Age15to24.Per, color = "Age 15-24"), size = 0.8) +
  geom_line(aes(y = Age25above.Per, color = "Age 25+"), size = 0.8) +
  geom_point(aes(y = Age15to24.Per, color = "Age 15-24"), size = 0.8, alpha = 0.5) +
  geom_point(aes(y = Age25above.Per, color = "Age 25+"), size = 0.8, alpha = 0.5) +
  labs(title = "Unemployment Rate by Age Group",
    x = "Year", y = "Colombia Unemployment Rate (%)", color = "Age Group") +
  scale_color_manual(values = c("Age 15-24" = "blue", "Age 25+" = "red")) +
  theme_gray(base_size = 12) +  
  theme(plot.title = element_text(size = 14),
    axis.title.x = element_text(size = 12), axis.title.y = element_text(size = 12),
    axis.text = element_text(size = 12), legend.position = "bottom") 

US1 <- ggplot(US, aes(x = Month)) +
  geom_line(aes(y = Age15to24.Per, color = "Age 15-24"), size = 0.8) +
  geom_line(aes(y = Age25above.Per, color = "Age 25+"), size = 0.8) +
  geom_point(aes(y = Age15to24.Per, color = "Age 15-24"), size = 0.8, alpha = 0.5) +
  geom_point(aes(y = Age25above.Per, color = "Age 25+"), size = 0.8, alpha = 0.5) +
  labs(title = "",
    x = "Year", y = "US Unemployment Rate (%)", color = "Age Group") +
  scale_color_manual(values = c("Age 15-24" = "blue", "Age 25+" = "red")) +
  theme_gray(base_size = 12) +  
  theme(plot.title = element_text(size = 14),
    axis.title.x = element_text(size = 12), axis.title.y = element_text(size = 12),
    axis.text = element_text(size = 12), legend.position = "bottom")

grid.arrange(PE1, US1, ncol = 2)

# Comparison of Unemployment Rate by Sex Group 
sPE1 <- ggplot(Colombia, aes(x = Month)) +
  geom_line(aes(y = Female.Per, color = "Female"), size = 0.8) +
  geom_line(aes(y = Male.Per, color = "Male"), size = 0.8) +
  geom_point(aes(y = Female.Per, color = "Female"), size = 0.8, alpha = 0.5) +
  geom_point(aes(y = Male.Per, color = "Male"), size = 0.8, alpha = 0.5) +
  labs(title = "Unemployment Rate by Sex Group",
    x = "Year", y = "Colombia Unemployment Rate (%)", color = "Sex Group") +
  scale_color_manual(values = c("Female" = "blue", "Male" = "red")) +
  theme_gray(base_size = 12) +  
  theme(plot.title = element_text(size = 14),
    axis.title.x = element_text(size = 12), axis.title.y = element_text(size = 12),
    axis.text = element_text(size = 12), legend.position = "bottom") 

sUS1 <- ggplot(US, aes(x = Month)) +
  geom_line(aes(y = Female.Per, color = "Female"), size = 0.8) +
  geom_line(aes(y = Male.Per, color = "Male"), size = 0.8) +
  geom_point(aes(y = Female.Per, color = "Female"), size = 0.8, alpha = 0.5) +
  geom_point(aes(y = Male.Per, color = "Male"), size = 0.8, alpha = 0.5) +
  labs(title = "",
    x = "Year", y = "US Unemployment Rate (%)", color = "Sex Group") +
  scale_color_manual(values = c("Female" = "blue", "Male" = "red")) +
  theme_gray(base_size = 12) +  
  theme(plot.title = element_text(size = 14),
    axis.title.x = element_text(size = 12), axis.title.y = element_text(size = 12),
    axis.text = element_text(size = 12), legend.position = "bottom") 

grid.arrange(sPE1, sUS1, ncol = 2)
```

# Additional graphs(Jisup)
# Innitial plot of Colombia as the perspective of the level of unemployment

```{r}
Colombia_unemployment <- Colombia

# Check for missing values
summary(Colombia_unemployment)
sum(is.na(Colombia_unemployment))


# Check regularity of time index
diff_date <- diff(Colombia_unemployment$Month)
table(diff_date)  # There's one missing month.
Colombia_unemployment$Month <- as.Date(Colombia_unemployment$Month)
Colombia_LF$Month <- as.Date(Colombia_LF$Month)


# Create full monthly sequence
full_month_seq <- data.frame(Month = seq.Date(from = min(Colombia_unemployment$Month),to = max(Colombia_unemployment$Month),
                                              by = "month"))

# Join with original to find missing months
missing_months <- full_month_seq %>%
  anti_join(Colombia_unemployment, by = "Month")
Colombia_LF <- full_month_seq %>%
  left_join(Colombia_LF, by = "Month")

print(missing_months)  # Missing month

# Fill in missing months with NA
Colombia_unemployment <- full_month_seq %>%
  left_join(Colombia_unemployment, by = "Month")


# Check for outliers visually
boxplot(Colombia_unemployment$Total.Thou,
        main = "Boxplot: Colombia AgeTotal.Thou",
        horizontal = TRUE,
        col = "lightblue")

boxplot(Colombia_unemployment$Total.Per,
        main = "Boxplot: Colombia Total Unemplyment Rate",
        horizontal = TRUE,
        col = "lightblue")

boxplot(Colombia_LF$Total.Thou,
        main = "Boxplot: Colombia AgeTotal.employment_Thou",
        horizontal = TRUE,
        col = "lightblue")



# Convert to time series
ts_Colombia_total_thou <- ts(Colombia_unemployment$Total.Thou,
               start = c(2001, 7), 
               frequency = 12)
ts_Colombia_total_per <- ts(Colombia_unemployment$Total.Per,
               start = c(2001, 7), 
               frequency = 12)
ts_Colombia_employment_thou <- ts(Colombia_LF$LF.Part,
               start = c(2001, 7), 
               frequency = 12)


# interpolation of NA
ts_Colombia_total_thou <- na_interpolation(ts_Colombia_total_thou, option = "linear")
ts_Colombia_total_per <- na_interpolation(ts_Colombia_total_per, option = "linear")
ts_Colombia_employment_thou <- na_interpolation(ts_Colombia_employment_thou, option = "linear")

sum(is.na(ts_Colombia_total_thou))
sum(is.na(ts_Colombia_total_per))
sum(is.na(ts_Colombia_employment_thou))

length(Colombia_unemployment$Month)           
length(ts_Colombia_total_thou)                
length(ts_Colombia_employment_thou)           
length(ts_Colombia_total_per)                 



# plotting the graph
Colombia_plot_data <- data.frame(
  Month = Colombia_unemployment$Month,
  Unemployment_Level = as.numeric(ts_Colombia_total_thou),
  Employment_Level = as.numeric(ts_Colombia_employment_thou),
  Unemployment_Rate_Percent = as.numeric(ts_Colombia_total_per)
)


p <- ggplot(Colombia_plot_data, aes(x = Month)) +
  geom_line(aes(y = Unemployment_Level*1.1, color = "Unemployment Level"), size = 1) +
  geom_line(aes(y = Employment_Level, color = "Employment Level"), size = 1) +
  geom_line(aes(y = Unemployment_Rate_Percent * 1000, color = "Unemployment Rate [%]"), size = 1) +
  scale_y_continuous(
    name = "Unemployment [Thousands]",
    limits = c(0, 30000),
    sec.axis = sec_axis(~./1000, name = "Unemployment Rate [%]", breaks = seq(0, 20, 5))
  ) +
  scale_color_manual(
    name = "Legend",
    values = c(
      "Unemployment Level" = "darkblue",
      "Employment Level" = "darkorange",
      "Unemployment Rate [%]" = "gray"
    )
  ) +
  labs(
    title = "Colombia: Employment, Unemployment Count and Rate Over Time",
    x = "Month"
  ) +
  theme_minimal()

print(p)

# Decompose the time series
decomp_ts_Colombia_total_thou <- decompose(ts_Colombia_total_thou)
plot(decomp_ts_Colombia_total_thou)
decomp_ts_Colombia_total_per <- decompose(ts_Colombia_total_per)
plot(decomp_ts_Colombia_total_per)
decomp_ts_Colombia_employment_thou <- decompose(ts_Colombia_employment_thou)
plot(decomp_ts_Colombia_employment_thou)

# ACF and PACF plots
par(mfrow = c(1, 2))
Acf(ts_Colombia_total_thou, lag.max = 40, main = "ACF: Colombia AgeTotal.Thou")
Pacf(ts_Colombia_total_thou, lag.max = 40, main = "PACF: Colombia AgeTotal.Thou")

par(mfrow = c(1, 2))
Acf(ts_Colombia_total_per, lag.max = 40, main = "ACF: Colombia AgeTotal.Thou")
Pacf(ts_Colombia_total_per, lag.max = 40, main = "PACF: Colombia AgeTotal.Thou")

par(mfrow = c(1, 2))
Acf(ts_Colombia_employment_thou, lag.max = 40, main = "ACF: Colombia AgeTotal.Thou")
Pacf(ts_Colombia_employment_thou, lag.max = 40, main = "PACF: Colombia AgeTotal.Thou")

# Grubbs test for outliers
grubbs.test(ts_Colombia_total_thou)
grubbs.test(ts_Colombia_total_per)
grubbs.test(ts_Colombia_employment_thou)
```

# Additional graphs(Jisup)
# Innitial plot of US as the perspective of the level of unemployment

```{r}
us_unemployment <- read_excel(
  path="./Data/Raw/Final Project_data_20250403_R.xlsx", sheet = "US_unemployment", col_names = TRUE)

# Check for missing values
summary(us_unemployment)
sum(is.na(us_unemployment))



# 1. Filtering after 2001
us_unemployment_filtered <- us_unemployment %>%
  filter(Month >= as.Date("2001-01-01"))

# 2. Plot
ggplot(us_unemployment_filtered, aes(x = Month)) +
  geom_line(aes(y = AgeTotal.Thou, color = "Unemployment Level"), size = 1) +
  geom_line(aes(y = Employment_level, color = "Employment Level"), size = 1) +
  geom_line(aes(y = AgeTotal.Per * 7500, color = "Unemployment Rate [%]"), size = 1) +
  scale_y_continuous(
    name = "Employment / Unemployment [Thousands]",
    limits = c(0, 150000), 
    sec.axis = sec_axis(~./7500, name = "Unemployment Rate [%]", breaks = seq(0, 20, 5))
  ) +
  scale_color_manual(
    name = "Legend",
    values = c(
      "Unemployment Level" = "darkblue",
      "Employment Level" = "darkorange",
      "Unemployment Rate [%]" = "gray"
    )
  ) +
  labs(
    title = "US: Employment, Unemployment Count and Rate Since 2001",
    x = "Month"
  ) +
  theme_minimal()



# Check regularity of time index
diff_date <- diff(us_unemployment$Month)
table(diff_date)  # There's one missing month.
us_unemployment$Month <- as.Date(us_unemployment$Month)


# Create full monthly sequence
full_month_seq <- data.frame(Month = seq.Date(from = min(us_unemployment$Month),to = max(us_unemployment$Month),
                                              by = "month"))

# Join with original to find missing months
missing_months <- full_month_seq %>%
  anti_join(us_unemployment, by = "Month")

print(missing_months) # no NAs

# Check for outliers visually
boxplot(us_unemployment$AgeTotal.Thou,
        main = "Boxplot: US AgeTotal.Thou",
        horizontal = TRUE,
        col = "lightblue")

boxplot(us_unemployment$AgeTotal.Per,
        main = "Boxplot: US AgeTotal.Thou",
        horizontal = TRUE,
        col = "lightblue")

boxplot(us_unemployment$Employment_level,
        main = "Boxplot: US AgeTotal.Thou",
        horizontal = TRUE,
        col = "lightblue")



# Convert to time series
ts_us_total_thou <- ts(us_unemployment$AgeTotal.Thou,
               start = c(2001, 4), 
               frequency = 12)
ts_us_total_per <- ts(us_unemployment$AgeTotal.Per,
               start = c(2001, 4), 
               frequency = 12)
ts_us_employment_thou <- ts(us_unemployment$Employment_level,
               start = c(2001, 4), 
               frequency = 12)

# Decompose the time series
decomp_ts_us_total_thou <- decompose(ts_us_total_thou)
plot(decomp_ts_us_total_thou)
decomp_ts_us_total_per <- decompose(ts_us_total_per)
plot(decomp_ts_us_total_per)
decomp_ts_us_employment_thou <- decompose(ts_us_employment_thou)
plot(decomp_ts_us_employment_thou)

# ACF and PACF plots
par(mfrow = c(1, 2))
Acf(ts_us_total_thou, lag.max = 40, main = "ACF: us AgeTotal.Thou")
Pacf(ts_us_total_thou, lag.max = 40, main = "PACF: us AgeTotal.Thou")

par(mfrow = c(1, 2))
Acf(ts_us_total_per, lag.max = 40, main = "ACF: us AgeTotal.Thou")
Pacf(ts_us_total_per, lag.max = 40, main = "PACF: us AgeTotal.Thou")

par(mfrow = c(1, 2))
Acf(ts_us_employment_thou, lag.max = 40, main = "ACF: us AgeTotal.Thou")
Pacf(ts_us_employment_thou, lag.max = 40, main = "PACF: us AgeTotal.Thou")

# Grubbs test for outliers
grubbs.test(ts_us_total_thou)
grubbs.test(ts_us_total_per)
grubbs.test(ts_us_employment_thou)
```

# Additional graphs(Jisup)
# Comparing Colombia and US(Population and GDP growth)
```{r}
# gdp_growth <- read_excel(
#   path="./Data/Raw/Final Project_data_20250403_R.xlsx", sheet = "US_GDP", col_names = TRUE)
# gdp_growth <- gdp_growth[244:nrow(gdp_growth), ]
# 
# 
# pop_growth <- read_excel(
#   path="./Data/Raw/Final Project_data_20250403_R.xlsx", sheet = "US_Population", col_names = TRUE)
# pop_growth <- pop_growth[80:nrow(pop_growth), ]
# 
# 
# # Check for missing values
# summary(gdp_growth)
# sum(is.na(gdp_growth))
# 
# summary(pop_growth)
# sum(is.na(pop_growth))


# 2. Plot
# ggplot(gdp_growth, aes(x = observation_date)) +
#   geom_line(aes(y = `Colombia GDP growth`, color = "Colombia GDP Growth"), size = 1) +
#   geom_line(aes(y = `US GDP growth`, color = "US GDP Growth"), size = 1) +
#   scale_color_manual(
#     name = "Legend",
#     values = c(
#       "Colombia GDP Growth" = "darkblue",
#       "US GDP Growth" = "darkorange"
#     )
#   ) +
#   labs(
#     title = "US & Colombia: Quarterly GDP Growth Since 2008",
#     x = "Date",
#     y = "GDP Growth (%)"
#   ) +
#   theme_minimal()
# 
# ggplot(pop_growth, aes(x = observation_date)) +
#   geom_line(aes(y = Colombia_pop_growth, color = "Colombia Population Growth"), size = 1) +
#   geom_line(aes(y = US_pop_growth, color = "US Population Growth"), size = 1) +
#   scale_color_manual(
#     name = "Legend",
#     values = c(
#       "Colombia Population Growth" = "darkblue",
#       "US Population Growth" = "darkorange"
#     )
#   ) +
#   labs(
#     title = "US & Colombia: Quarterly Population Growth Since 2008",
#     x = "Date",
#     y = "Population Growth (%)"
#   ) +
#   theme_minimal()

```



### Summary Statistics 
```{r Colombia summary stats table}
## Colombia
# Generate Summary Statistics
summary_table <- Colombia %>%
  select(-Month) %>%  # Exclude Month column
  summarise(across(where(is.numeric), 
                   list(Mean = ~ mean(.x),
                        SD = ~ sd(.x),
                        Min = ~ min(.x),
                        Max = ~ max(.x),
                        N = ~ sum(!is.na(.x))))) %>%
  pivot_longer(everything(), names_to = c("Variable", ".value"), names_sep = "_") %>% 
  gt() %>%
  tab_header(title = "Summary Statistics of Unemployment Data in Colombia",
    subtitle = "Monthly Data (2001-2024)") %>%
  fmt_number(columns = 2:6, decimals = 2) %>%
  cols_label(Variable = "Indicator", Mean = "Mean", SD = "Standard Deviation",
    Min = "Min", Max = "Max", N = "Observations") %>%
  tab_options(table.font.size = px(14),
    heading.title.font.size = px(18), heading.subtitle.font.size = px(14))

print(summary_table)   # During knitting, this should not be excuted.

summary_table <- Colombia %>%
  select(-Month) %>%
  summarise(across(where(is.numeric), 
                   list(Mean = ~ mean(.x),
                        SD = ~ sd(.x),
                        Min = ~ min(.x),
                        Max = ~ max(.x),
                        N = ~ sum(!is.na(.x))))) %>%
  pivot_longer(everything(), names_to = c("Variable", ".value"), names_sep = "_")

knitr::kable(summary_table, digits = 2, caption = "Summary Statistics of Unemployment Data in Colombia")


# Check outliers 
outlier(Colombia) 
grubbs.test(Colombia$Age15to24.Thou) # This is an outlier. 
grubbs.test(Colombia$Age25above.Thou) # This is an outlier. 
grubbs.test(Colombia$Age15to24.Per) 
grubbs.test(Colombia$Age25above.Per)
grubbs.test(Colombia$Female.Thou) # This is an outlier. 
grubbs.test(Colombia$Male.Thou) # This is an outlier. 
grubbs.test(Colombia$Female.Per)
grubbs.test(Colombia$Male.Per)
grubbs.test(Colombia$Total.Thou) # This is an outlier. 
grubbs.test(Colombia$Total.Per)

# Check the box plot for total unemployment (for PPT)
boxplot(Colombia$Total.Per,
        main = "Boxplot: Colombia Unemployment Rate (%)",
        horizontal = TRUE, 
        col = "lightblue")

# Plotting for total unemployment, ACF, PACF (for PPT)
ts_Colombia_ppt <- ts(Colombia$Total.Per, start = c(2001, 1), frequency = 12)
p1 <- autoplot(ts_Colombia_ppt) +
  ggtitle("Colombia") +
  ylab("Unemployment Rate (%)") +
  xlab("Time")

p2 <- ggAcf(ts_Colombia_ppt, lag.max = 40) +
  ggtitle("ACF")

p3 <- ggPacf(ts_Colombia_ppt, lag.max = 40) +
  ggtitle("PACF")

grid.arrange(p1, p2, p3, ncol = 3)


# Find the row where Total.Per is the outlier (10.8%)
outlier_row <- Colombia %>%
  filter(Total.Per == max(Total.Per, na.rm = TRUE))

print(outlier_row) # Highest Value is 2005-02-01. 
```

## Time Series Analysis for Colombia 
### Step 1: Transform into time series and set training and testing windows for Colombia

```{r}
# Transform into time series
# 201001~202412
# Transform into time series
ts_Colombia_total_per


ts_Colombia <- ts(Colombia[,2:11],
              start=c(year(Colombia$Month[1]), month(Colombia$Month[1])),
              frequency = 12)

# Set the period
nobs = nrow(Colombia)
n_for = 12

# Create a subset for training purpose 
ts_Colombia_train <- ts(Colombia[97:(nobs-n_for),2:11],
                    start=c(2010, 1),
                    frequency = 12)
#ts_Colombia_total <- window(ts_Colombia_total_per, start = c(2010, 1), end = c(2023, 12))

head(ts_Colombia_train)
tail(ts_Colombia_train)

# Create a subset for testing purpose(2024)
start_row = nobs - n_for + 1
ts_Colombia_test <- ts(Colombia[(nobs-n_for+1):nobs,2:11],
                   start=c(year(Colombia$Month[start_row]),
                           month(Colombia$Month[start_row])), frequency = 12)

#ts_Colombia_test <- window(ts_Colombia_total_per, start = c(2024, 1), end = c(2024, 12))

head(ts_Colombia_test)
tail(ts_Colombia_test)

# Plots 
train <- autoplot(ts_Colombia_train[,"Total.Per"]) + ylab("Unemployment Rate (%)") +
  ggtitle("Training Window")
test <- autoplot(ts_Colombia_test[,"Total.Per"]) + ylab("Unemployment Rate (%)") +
  ggtitle("Testing Window")
grid.arrange(train, test, ncol = 2)

par(mfrow=c(1,2))
  Acf(ts_Colombia_train[,"Total.Per"], lag=40, plot = TRUE, main = "")
  Pacf(ts_Colombia_train[,"Total.Per"], lag=40, plot = TRUE, main = "")

```

### Step 2: Decompose the time series for Colombia
```{r total.per decompose of Colombia, warning=FALSE, message=FALSE}
# Decompose 
decom_totalper_train <- decompose(ts_Colombia_train[,"Total.Per"])
plot(decom_totalper_train)

# Deseason 
deseas_totalper_train <- seasadj(decom_totalper_train)  
plot(deseas_totalper_train)

# Run the tests on deseasoned series
print(adf.test(deseas_totalper_train, alternative = "stationary")) # It is unit root. 
summary(MannKendall(deseas_totalper_train)) # It has a decreasing trend.

# Run the tests on original series 
print(adf.test(ts_Colombia_train[,"Total.Per"], alternative = "stationary")) # It is stationary. 
summary(SeasonalMannKendall(ts_Colombia_train[,"Total.Per"])) 
summary(smk.test(ts_Colombia_train[,"Total.Per"])) # It has seasonality. 

# Check for any differencing needed 
print(ndiffs(ts_Colombia_train[,"Total.Per"]))
print(ndiffs(deseas_totalper_train))
```

### Step 3: Test Time Series Models for Colombia
```{r time series models Colombia, warning=FALSE, message=FALSE}
# Seasonal Naive Model 
SNAIVE_deseas_totalper <- snaive(ts_Colombia_train[,"Total.Per"], h=n_for)
autoplot(SNAIVE_deseas_totalper)
checkresiduals(SNAIVE_deseas_totalper)

# Simple Moving Average Model
SMA_deseas_totalper <- smooth::sma(y = deseas_totalper_train, h=n_for, 
                                   holdout = FALSE, silent = FALSE) 
summary(SMA_deseas_totalper)
autoplot(SMA_deseas_totalper$fitted)
checkresiduals(SMA_deseas_totalper) # Residuals are not iid.

# Simple Exponential Smoothing Model
SES_deseas_totalper = ses(y = deseas_totalper_train, h=n_for, 
                          holdout = FALSE, silent = FALSE)  
summary(SES_deseas_totalper)
autoplot(SES_deseas_totalper)
checkresiduals(SES_deseas_totalper) # Residuals are not iid. 

# SARIMA Model
SARIMA_totalper <- auto.arima(ts_Colombia_train[,"Total.Per"])
print(SARIMA_totalper)

SARIMA_forecast_totalper <- forecast(object = SARIMA_totalper, h=n_for)
autoplot(SARIMA_forecast_totalper)
checkresiduals(SARIMA_forecast_totalper) # Residuals are iid.

# Deaseasoned ARIMA Model
ARIMA_totalper <- auto.arima(deseas_totalper_train, max.D = 0, 
                             max.P = 0, max.Q = 0)
print(ARIMA_totalper)

ARIMA_forecast_totalper <- forecast(object = ARIMA_totalper, h=n_for)
autoplot(ARIMA_forecast_totalper)
checkresiduals(ARIMA_forecast_totalper) # Residuals are iid.
 
# STL + ETS Model
ETS_totalper <-  stlf(ts_Colombia_train[,"Total.Per"],h=n_for)
autoplot(ETS_totalper) 
checkresiduals(ETS_totalper) # Residuals are not iid. 

# ARIMA + FOURIER Model
ARIMA_Four_fit_totalper <- auto.arima(ts_Colombia_train[,"Total.Per"], 
                             seasonal=FALSE, lambda=0,
                             xreg=fourier(ts_Colombia_train[,"Total.Per"], 
                                          K=3))

ARIMA_Four_for_totalper <- forecast(ARIMA_Four_fit_totalper,
                           xreg=fourier(ts_Colombia_train[,"Total.Per"],
                                        K=3, h=n_for),
                           h=n_for) 

autoplot(ARIMA_Four_for_totalper)
checkresiduals(ARIMA_Four_for_totalper) # Better fit

# TBATS Model 
TBATS_fit_totalper <- tbats(ts_Colombia_train[,"Total.Per"])
TBATS_for_totalper <- forecast(TBATS_fit_totalper, h = n_for)
autoplot(TBATS_for_totalper) 
checkresiduals(TBATS_fit_totalper) # Better fit

# Neural Network Model 
NN_fit_totalper <- nnetar(ts_Colombia_train[,"Total.Per"],
                          p=3, P=0,
                          xreg=fourier(ts_Colombia_train[,"Total.Per"], K=3))

NN_for_totalper <- forecast(NN_fit_totalper, 
                   h=n_for,
                   xreg=fourier(ts_Colombia_train[,"Total.Per"], 
                                          K=3,h=n_for))

autoplot(NN_for_totalper)
checkresiduals(NN_fit_totalper) # Residuals are not iid.

## State Space Exponential Smoothing Model
SSES_seas_totalper <- es(ts_Colombia_train[,"Total.Per"],
                         model="ZZZ", h=n_for, holdout=FALSE)

plot(SSES_seas_totalper$fitted, type = "l", col = "blue", main = "Fitted Values from SSES Model")
lines(SSES_seas_totalper$forecast, col = "red")
checkresiduals(SSES_seas_totalper) # Residuals are not iid.

## State Space with BSM Model
SS_seas_totalper <- StructTS(ts_Colombia_train[,"Total.Per"],
                    type="BSM",fixed=c(0.01,0.001,0.1,NA)) 

SS_for_totalper <- forecast(SS_seas_totalper,h=n_for)

plot(SS_for_totalper)
checkresiduals(SS_seas_totalper) # Residuals are not iid. 
```

### Step 4: Performance check for Colombia
```{r accuracy Colombia, warning=FALSE, message=FALSE}
# Check accuracy of the models
SANIVE_tpscores <- accuracy(SNAIVE_deseas_totalper$mean,ts_Colombia_test[,"Total.Per"])  
SMA_tpscores <- accuracy(SMA_deseas_totalper$forecast,ts_Colombia_test[,"Total.Per"])  
SES_tpscores <- accuracy(SES_deseas_totalper$mean,ts_Colombia_test[,"Total.Per"])
SARIMA_tpscores <- accuracy(SARIMA_forecast_totalper$mean,ts_Colombia_test[,"Total.Per"])
ARIMA_tpscores <- accuracy(ARIMA_forecast_totalper$mean,ts_Colombia_test[,"Total.Per"])
ETS_tpscores <- accuracy(ETS_totalper$mean,ts_Colombia_test[,"Total.Per"])
ARIMA_Four_tpscores <- accuracy(ARIMA_Four_for_totalper$mean,ts_Colombia_test[,"Total.Per"])
TBATS_tpscores <- accuracy(TBATS_for_totalper$mean,ts_Colombia_test[,"Total.Per"])
NN_tpscores <- accuracy(NN_for_totalper$mean,ts_Colombia_test[,"Total.Per"])
SSES_tpscores <- accuracy(SSES_seas_totalper$forecast,ts_Colombia_test[,"Total.Per"])
SS_tpscores <- accuracy(SS_for_totalper$mean,ts_Colombia_test[,"Total.Per"])

# Compare the matrix 
tpscores <- as.data.frame(rbind(SANIVE_tpscores, SMA_tpscores, 
                                SES_tpscores, SARIMA_tpscores, ARIMA_tpscores, 
                                ETS_tpscores, ARIMA_Four_tpscores, TBATS_tpscores, 
                                NN_tpscores, SSES_tpscores, SS_tpscores)) %>%
  mutate(Average = rowMeans(select(., RMSE, MAPE), na.rm = TRUE))

row.names(tpscores) <- c("SNAIVE", "SMA", "SES", "SARIMA", "ARIMA",
                       "ETS", "ARIMA_FOURIER", "TBATS", "NNETAR",
                       "SSES", "BSM")

# Choose model with lowest error
best_model_index_tp <- which.min(tpscores[,"Average"])
cat("The best model by Average(RMSE$MAPE) is:", row.names(tpscores[best_model_index_tp,]))  

# Create Tables 
kbl(tpscores, 
      caption = "Forecast Accuracy for Unemployment Rate (%) Data(Average Based on RMSE and MAPE Only)",
      digits = array(5,ncol(tpscores))) %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  #highlight model with lowest RMSE
  kable_styling(latex_options="striped", stripe_index = which.min(tpscores[,"Average"]))

# Plot everything together

autoplot(ts_Colombia_test[,"Total.Per"], size=0.8) +
  ylab("Unemployment Rate [%]") +
  autolayer(SNAIVE_deseas_totalper, PI = FALSE, series = "SNAIVE", size = 0.8) + 
  autolayer(SES_deseas_totalper, PI = FALSE, series = "SES", size = 0.8) +
  autolayer(SARIMA_forecast_totalper, PI = FALSE, series = "SARIMA", size = 0.8) +
  autolayer(ARIMA_forecast_totalper, PI = FALSE, series = "ARIMA", size = 0.8) +
  autolayer(ETS_totalper, PI = FALSE, series = "ETS", size = 0.8) +
  autolayer(ARIMA_Four_for_totalper, PI = FALSE, series = "ARIMA_FOURIER", size = 0.8) +
  autolayer(TBATS_for_totalper, PI = FALSE, series = "TBATS", size = 0.8) +
  autolayer(NN_for_totalper, PI = FALSE, series = "NNETAR", size = 0.8) +
  autolayer(SS_for_totalper, PI = FALSE, series = "BSM", size = 0.8) +
  scale_x_continuous(breaks = seq(2024 + 1/12, 2024 + 12/12, by = 1/12),
                     labels = month.abb) +
  guides(colour = guide_legend(title = "Forecast"))

```

### Step 5: Forecast for 2025 with the best three models for Colombia
```{r forecast 2025, warning=FALSE, message=FALSE}


n_full = 12

# Create the time series to retain full data set
ts_Colombia_fulltrain <- ts(Colombia[97:276,6],
              start=c(2010, 1),
              frequency = 12)

# Neural Network Model _1st
NN_fit_totalper <- nnetar(ts_Colombia_fulltrain[,"Total.Per"],
                          p=3, P=0,
                          xreg=fourier(ts_Colombia_fulltrain[,"Total.Per"], K=3))

NN_for_totalper <- forecast(NN_fit_totalper, 
                   h=n_for,
                   xreg=fourier(ts_Colombia_fulltrain[,"Total.Per"], 
                                          K=3,h=n_for))

autoplot(NN_for_totalper)
checkresiduals(NN_fit_totalper) # Residuals are not iid.

# ARIMA + FOURIER Model_2nd
ARIMA_Four_fit_totalper <- auto.arima(ts_Colombia_fulltrain[,"Total.Per"], 
                             seasonal=FALSE, lambda=0,
                             xreg=fourier(ts_Colombia_fulltrain[,"Total.Per"], 
                                          K=3))

ARIMA_Four_for_totalper <- forecast(ARIMA_Four_fit_totalper,
                           xreg=fourier(ts_Colombia_fulltrain[,"Total.Per"],
                                        K=3, h=n_for),
                           h=n_for) 

autoplot(ARIMA_Four_for_totalper)
checkresiduals(ARIMA_Four_for_totalper) # Better fit


# SARIMA Model_3nd
SARIMA_totalper <- auto.arima(ts_Colombia_fulltrain[,"Total.Per"])
print(SARIMA_totalper)

SARIMA_forecast_totalper <- forecast(object = SARIMA_totalper, h=n_for)
autoplot(SARIMA_forecast_totalper)
checkresiduals(SARIMA_forecast_totalper) # Residuals are iid.


# Plot 3 models together 
autoplot(ts_Colombia_fulltrain) +
  autolayer(NN_for_totalper, series="Nueral Network with Fourier",PI=FALSE)+
  autolayer(ARIMA_Four_for_totalper, series="ARIMA with Fourier",PI=FALSE)+
  autolayer(SARIMA_forecast_totalper, series="SARIMA",PI=FALSE)+
  ylab("Unemployment Rate (%)") + 
  ggtitle("Forecasted Unemployment Rate (%) in Colombia")



```

# Step 6: The average of 3 forecasts
```{r}

# --- 1. Extract predicted means ---
bsm_fc <- as.numeric(NN_for_totalper$mean)
sses_fc <- as.numeric(ARIMA_Four_for_totalper$mean)
ets_fc <- as.numeric(SARIMA_forecast_totalper$mean)

# Compute average forecast
avg_fc <- (bsm_fc + sses_fc + ets_fc) / 3

# --- 2. Create forecast date index ---
# 마지막 월 기준 + 1개월
start_month <- as.Date("2025-01-01")  # 또는 Colombia$Month의 마지막 날짜 이후
forecast_dates <- seq(from = start_month, by = "month", length.out = n_full)

# --- 3. Create forecast dataframe ---
forecast_df <- tibble(
  Month = forecast_dates,
  BSM = bsm_fc,
  SSES = sses_fc,
  ETS = ets_fc,
  Avg_Forecast = avg_fc
)

# --- 4. Export to Excel ---
write.xlsx(forecast_df, "./Forecast_Average_Colombia.xlsx")

# --- 5. Extract actuals from 2020 ---
Colombia_actual_2020on <- Colombia %>%
  filter(Month >= as.Date("2010-01-01")) %>%
  select(Month, Actual = Total.Per)

# Combine with forecast
plot_df <- bind_rows(
  tibble(Month = Colombia_actual_2020on$Month,
         Value = Colombia_actual_2020on$Actual,
         Type = "Actual"),
  tibble(Month = forecast_dates,
         Value = avg_fc,
         Type = "Avg Forecast")
)

# --- 6. Plot actuals + forecast ---
ggplot(plot_df, aes(x = Month, y = Value, color = Type, group = 1)) +
  geom_line(size = 1) +
  scale_color_manual(values = c("Actual" = "black", "Avg Forecast" = "steelblue")) +
  labs(
    title = "Average of Top 3 Forecast Result(Colombia)",
    y = "Unemployment Rate (%)",
    x = "Month",
    color = "Legend"
  ) +
  theme_minimal()

# --- 7. Extract actuals from 2022 ---
Colombia_actual_2022on <- Colombia %>%
  filter(Month >= as.Date("2022-01-01")) %>%
  select(Month, Actual = Total.Per)

# Combine with forecast
plot_df <- bind_rows(
  tibble(Month = Colombia_actual_2022on$Month,
         Value = Colombia_actual_2022on$Actual,
         Type = "Actual"),
  tibble(Month = forecast_dates,
         Value = avg_fc,
         Type = "Avg Forecast")
)

# --- 8. Plot actuals + forecast ---
ggplot(plot_df, aes(x = Month, y = Value, color = Type, group = 1)) +
  geom_line(size = 1) +
  scale_color_manual(values = c("Actual" = "black", "Avg Forecast" = "steelblue")) +
  labs(
    title = "Average of Top 3 Forecast Result(Colombia)",
    y = "Unemployment Rate (%)",
    x = "Month",
    color = "Legend"
  ) +
  theme_minimal()

```

