---
title: 'Forecasting Unemployment Trends: A Comparative Time Series Analysis of Colombia
  and the United States'
author: "Aye Nyein Thu, Mazhar Bhuyan, Yuqi Yang, Jisup Kwak"
date: "2025-03-24"
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---

## Setting R code chunk options (To Remove)
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  error = TRUE,   
  results = 'markup',
  tidy.opts = list(width.cutoff = 80),
  tidy = FALSE
)
```

## Loading packages and initializing (To Remove)
```{r packages, warning=FALSE, message=FALSE}
# Load required packages
library(readxl)
library(openxlsx)
library(writexl)
library(dplyr)
library(lubridate)
library(ggplot2)
library(cowplot)
library(forecast)  
library(Kendall)
library(tseries)
library(outliers)
library(tidyverse)
library(smooth)
library(trend)
library(kableExtra)
library(tidyr)
library(gt) #install.packages("gt")
library(gridExtra) #install.packages("gridExtra")
library(zoo)
library(imputeTS)


# Check working directory
getwd()
```

## Importing and Wrangling Data (To remove)
```{r importing data, results='hide', warning=FALSE, message=FALSE}
## Raw Data Set: Unemployment Rate by Age (Thousands)
# Import data set
UEAge.Thou <- read_excel(
  path="./Data/Raw/UE_Age(Thousands).xlsx", sheet = "Sheet1", col_names = TRUE)

# Format data set
UEAge.Thou_Processed <- UEAge.Thou %>%
  mutate(
    Month = ym(sub("M", "-", Month)), 
    Age15to24.Thou = as.numeric(`15-24`), 
    Age25above.Thou = as.numeric(`25+`),   
    AgeTotal.Thou = as.numeric(`15+`)) %>% 
  rename(Country="Reference area") %>% 
  select(Country,Month,Age15to24.Thou, Age25above.Thou, AgeTotal.Thou) %>% 
  arrange(Country, Month)

## Raw Data Set: Unemployment Rate by Age (%)
# Import data set
UEAge.Per <- read_excel(
  path="./Data/Raw/UE_Age(%).xlsx", sheet = "Sheet1", col_names = TRUE)

# Format data set
UEAge.Per_Processed <- UEAge.Per %>%
  mutate(
    Month = ym(sub("M", "-", Month)), 
    Age15to24.Per = as.numeric(`15-24`), 
    Age25above.Per = as.numeric(`25+`),   
    AgeTotal.Per = as.numeric(`15+`)) %>% 
  rename(Country="Reference area") %>% 
  select(Country,Month,Age15to24.Per, Age25above.Per, AgeTotal.Per) %>% 
  arrange(Country, Month)

## Raw Data Set: Unemployment Rate by Gender (Thousands)
# Import data set
UEGender.Thou <- read_excel(
  path="./Data/Raw/UE_Gender(Thousands).xlsx", sheet = "Sheet1", col_names = TRUE)

# Format data set
UEGender.Thou_Processed <- UEGender.Thou %>%
  mutate(
    Month = ym(sub("M", "-", Month)), 
    Female.Thou = as.numeric(Female), 
    Male.Thou = as.numeric(Male),   
    Total.Thou = as.numeric(Total)) %>% 
  rename(Country="Reference area") %>% 
  select(Country,Month,Female.Thou, Male.Thou, Total.Thou) %>% 
  arrange(Country, Month)

## Raw Data Set: Unemployment Rate by Gender (%)
# Import data set
UEGender.Per <- read_excel(
  path="./Data/Raw/UE_Gender(%).xlsx", sheet = "Sheet1", col_names = TRUE)

# Format data set
UEGender.Per_Processed <- UEGender.Per %>%
  mutate(
    Month = ym(sub("M", "-", Month)), 
    Female.Per = as.numeric(Female), 
    Male.Per = as.numeric(Male),   
    Total.Per = as.numeric(Total)) %>% 
  rename(Country="Reference area") %>% 
  select(Country,Month,Female.Per, Male.Per, Total.Per) %>% 
  arrange(Country, Month)
```

## To Remove 
```{r data wrangling US, warning=FALSE, message=FALSE}
# Combine all processed data sets 
UE_Countries <- UEAge.Thou_Processed %>% 
  left_join(UEAge.Per_Processed, by=c("Country", "Month")) %>% 
  left_join(UEGender.Thou_Processed, by=c("Country", "Month")) %>% 
  left_join(UEGender.Per_Processed, by=c("Country", "Month")) 

# Extract Colombia Data
Colombia <- UE_Countries %>% 
  filter(Country == "Colombia") %>% 
  select(-Country, AgeTotal.Per, AgeTotal.Thou) %>% 
  select(Month, Age15to24.Per, Age25above.Per, Female.Per, Male.Per,
         Total.Per, Age15to24.Thou, Age25above.Thou, Female.Thou, Male.Thou,
         Total.Thou) 


# Check Missing Value 
sum(is.na(Colombia))

# Extract US Data 
US <- UE_Countries %>% 
  filter(Country == "United States of America",
         Month >= as.Date("2001-01-01") & Month <= as.Date("2024-12-01")) %>% 
  select(-Country, AgeTotal.Per, AgeTotal.Thou) %>% 
  select(Month, Age15to24.Per, Age25above.Per, Female.Per, Male.Per,
         Total.Per, Age15to24.Thou, Age25above.Thou, Female.Thou, Male.Thou,
         Total.Thou) 

# Check Missing Value 
sum(is.na(US))
```

### Summary Statistics (US)
```{r US summary stats table}
## US
# Generate Summary Statistics
summary_tableUS <- US %>%
  select(-Month) %>%  # Exclude Month column
  summarise(across(where(is.numeric), 
                   list(Mean = ~ mean(.x),
                        SD = ~ sd(.x),
                        Min = ~ min(.x),
                        Max = ~ max(.x),
                        N = ~ sum(!is.na(.x))))) %>%
  pivot_longer(everything(), names_to = c("Variable", ".value"), names_sep = "_") %>% 
  gt() %>%
  tab_header(title = "Summary Statistics of Unemployment Data in US",
    subtitle = "Monthly Data (2001-2024)") %>%
  fmt_number(columns = 2:6, decimals = 2) %>%
  cols_label(Variable = "Indicator", Mean = "Mean", SD = "Standard Deviation",
    Min = "Min", Max = "Max", N = "Observations") %>%
  tab_options(table.font.size = px(14),
    heading.title.font.size = px(18), heading.subtitle.font.size = px(14))

print(summary_tableUS) # During knitting, this should not be executed.

summary_tableUS_simple <- US %>%
  select(-Month) %>%  
  summarise(across(where(is.numeric), 
                   list(Mean = ~ mean(.x),
                        SD = ~ sd(.x),
                        Min = ~ min(.x),
                        Max = ~ max(.x),
                        N = ~ sum(!is.na(.x))))) %>%
  pivot_longer(everything(), names_to = c("Variable", ".value"), names_sep = "_")

knitr::kable(summary_tableUS_simple, digits = 2, caption = "Summary Statistics of Unemployment Data in US")


# Check outliers 
outlier(US) 
grubbs.test(US$Age15to24.Thou) 
grubbs.test(US$Age25above.Thou) # This is an outlier. 
grubbs.test(US$Age15to24.Per) # This is an outlier. 
grubbs.test(US$Age25above.Per) # This is an outlier. 
grubbs.test(US$Female.Thou) # This is an outlier. 
grubbs.test(US$Male.Thou) 
grubbs.test(US$Female.Per) # This is an outlier. 
grubbs.test(US$Male.Per)
grubbs.test(US$Total.Thou) # This is an outlier. 
grubbs.test(US$Total.Per) # This is an outlier. 

# Check the box plot for total unemployment 
boxplot(US$Total.Per,
        main = "Boxplot: US Unemployment Rate (%)",
        horizontal = TRUE, 
        col = "lightblue")

# Find the row where Total.Per is the outlier (14.4%)
outlier_rowUS <- US %>%
  filter(Total.Per == max(Total.Per, na.rm = TRUE))

print(outlier_rowUS) # Outlier is 2020-04-01. 
```

## Time Series Analysis for US (Original Series)
### Step 1: Transform into time series and set training and testing windows for US (Original)
```{r time series}
# Transform into time series
ts_US <- ts(US[,2:11],
              start=c(year(US$Month[1]), month(US$Month[1])),
              frequency = 12)

# Set the period
nobsUS = nrow(US)
n_forUS = 12

# Create a subset for training purpose 
ts_US_train <- ts(US[1:(nobsUS-n_forUS),2:11],
                    start=c(year(US$Month[1]), month(US$Month[1])),
                    frequency = 12)

# Create a subset for testing purpose
start_rowUS = nobsUS - n_forUS + 1
ts_US_test <- ts(US[(nobsUS - n_forUS + 1):nobsUS,2:11],
                   start=c(year(US$Month[start_rowUS]),
                           month(US$Month[start_rowUS])), frequency = 12)

# Plots 
trainUS <- autoplot(ts_US_train[,"Total.Per"]) + ylab("Unemployment Rate (%)") +
  ggtitle("Training Window")
testUS <- autoplot(ts_US_test[,"Total.Per"]) + ylab("Unemployment Rate (%)") +
  ggtitle("Testing Window")
grid.arrange(trainUS, testUS, ncol = 2)

par(mfrow=c(1,2))
Acf(ts_US_train[,"Total.Per"], lag=40, plot = TRUE, main = "")
Pacf(ts_US_train[,"Total.Per"], lag=40, plot = TRUE, main = "")
par(mfrow=c(1,1))
```

### Step 2: Decompose the time series for US (Original)
```{r total.per decompose of US, warning=FALSE, message=FALSE}
# Decompose 
decom_totalper_trainUS <- decompose(ts_US_train[,"Total.Per"])
plot(decom_totalper_trainUS)

# Deseason 
deseas_totalper_trainUS <- seasadj(decom_totalper_trainUS)  
plot(deseas_totalper_trainUS)

# Run the tests on deseasoned series
print(adf.test(deseas_totalper_trainUS, alternative = "stationary")) # It is stationary. 
summary(MannKendall(deseas_totalper_trainUS)) # It has a decreasing trend.

# Run the tests on original series 
print(adf.test(ts_US_train[,"Total.Per"], alternative = "stationary")) # It is stationary. 
summary(SeasonalMannKendall(ts_US_train[,"Total.Per"])) 
summary(smk.test(ts_US_train[,"Total.Per"])) # It has seasonality. 

# Check for any differencing needed 
print(ndiffs(ts_US_train[,"Total.Per"]))
print(ndiffs(deseas_totalper_trainUS))
```

### Step 3: Test Time Series Models for US (Original)
```{r time series models, warning=FALSE, message=FALSE}
# Seasonal Naive Model 
SNAIVE_deseas_totalperUS <- snaive(ts_US_train[,"Total.Per"], h=n_forUS)
autoplot(SNAIVE_deseas_totalperUS)
checkresiduals(SNAIVE_deseas_totalperUS) # Residuals are not iid. 

# Simple Moving Average Model
SMA_deseas_totalperUS <- smooth::sma(y = deseas_totalper_trainUS, h=n_forUS, 
                                     holdout = FALSE, silent = FALSE) 
summary(SMA_deseas_totalperUS)
checkresiduals(SMA_deseas_totalperUS) # Residuals are iid. 

# Simple Exponential Smoothing Model
SES_deseas_totalperUS = ses( y = deseas_totalper_trainUS, h=n_forUS, 
                             holdout = FALSE, silent = FALSE)  
summary(SES_deseas_totalperUS)
autoplot(SES_deseas_totalperUS)
checkresiduals(SES_deseas_totalperUS) # Residuals are iid. 

# SARIMA Model
SARIMA_totalperUS <- auto.arima(ts_US_train[,"Total.Per"])
print(SARIMA_totalperUS)

SARIMA_forecast_totalperUS <- forecast(object = SARIMA_totalperUS, h=n_forUS)
autoplot(SARIMA_forecast_totalperUS)
checkresiduals(SARIMA_forecast_totalperUS) # Residuals are iid.

# Deaseasoned ARIMA Model
ARIMA_totalperUS <- auto.arima(deseas_totalper_trainUS, max.D = 0, 
                               max.P = 0, max.Q = 0)
print(ARIMA_totalperUS)

ARIMA_forecast_totalperUS <- forecast(object = ARIMA_totalperUS, h=n_forUS)
autoplot(ARIMA_forecast_totalperUS)
checkresiduals(ARIMA_forecast_totalperUS) # Residuals are iid.
 
# STL + ETS Model
ETS_totalperUS <-  stlf(ts_US_train[,"Total.Per"],h=n_forUS)
autoplot(ETS_totalperUS) 
checkresiduals(ETS_totalperUS) # Residuals are iid. 

# ARIMA + FOURIER Model
ARIMA_Four_fit_totalperUS <- auto.arima(ts_US_train[,"Total.Per"], 
                             seasonal=FALSE, lambda=0,
                             xreg=fourier(ts_US_train[,"Total.Per"], 
                                          K=3))

ARIMA_Four_for_totalperUS <- forecast(ARIMA_Four_fit_totalperUS,
                           xreg=fourier(ts_US_train[,"Total.Per"],
                                        K=3, h=n_forUS),
                           h=n_forUS) 

autoplot(ARIMA_Four_for_totalperUS)
checkresiduals(ARIMA_Four_for_totalperUS) # Residuals are iid. 

# TBATS Model 
TBATS_fit_totalperUS <- tbats(ts_US_train[,"Total.Per"])
TBATS_for_totalperUS <- forecast(TBATS_fit_totalperUS, h = n_forUS)
autoplot(TBATS_for_totalperUS) 
checkresiduals(TBATS_fit_totalperUS) # Residuals are iid. 

# Neural Network Model 
NN_fit_totalperUS <- nnetar(ts_US_train[,"Total.Per"],
                 p=3, P=0,
                 xreg=fourier(ts_US_train[,"Total.Per"], K=3))

NN_for_totalperUS <- forecast(NN_fit_totalperUS, 
                   h=n_forUS,
                   xreg=fourier(ts_US_train[,"Total.Per"], 
                                          K=3,h=n_forUS))

autoplot(NN_for_totalperUS)
checkresiduals(NN_fit_totalperUS) # Residuals are iid. 

## State Space Exponential Smoothing Model
SSES_seas_totalperUS <- es(ts_US_train[,"Total.Per"],
                         model="ZZZ", h=n_forUS, holdout=FALSE)
checkresiduals(SSES_seas_totalperUS) # Residuals are iid.

## State Space with BSM Model
SS_seas_totalperUS <- StructTS(ts_US_train[,"Total.Per"],
                    type="BSM",fixed=c(0.01,0.001,0.1,NA)) 

SS_for_totalperUS <- forecast(SS_seas_totalperUS,h=n_forUS)

plot(SS_for_totalperUS)
checkresiduals(SS_seas_totalperUS) # Residuals are not iid. 
```

### Step 4: Performance check for US (Original)
```{r accuracy US, warning=FALSE, message=FALSE}
# Check accuracy of the models
SANIVE_tpscoresUS <- accuracy(SNAIVE_deseas_totalperUS$mean,ts_US_test[,"Total.Per"])  
SMA_tpscoresUS <- accuracy(SMA_deseas_totalperUS$forecast,ts_US_test[,"Total.Per"])  
SES_tpscoresUS <- accuracy(SES_deseas_totalperUS$mean,ts_US_test[,"Total.Per"])
SARIMA_tpscoresUS <- accuracy(SARIMA_forecast_totalperUS$mean,ts_US_test[,"Total.Per"])
ARIMA_tpscoresUS <- accuracy(ARIMA_forecast_totalperUS$mean,ts_US_test[,"Total.Per"])
ETS_tpscoresUS <- accuracy(ETS_totalperUS$mean,ts_US_test[,"Total.Per"])
ARIMA_Four_tpscoresUS <- accuracy(ARIMA_Four_for_totalperUS$mean,ts_US_test[,"Total.Per"])
TBATS_tpscoresUS <- accuracy(TBATS_for_totalperUS$mean,ts_US_test[,"Total.Per"])
NN_tpscoresUS <- accuracy(NN_for_totalperUS$mean,ts_US_test[,"Total.Per"])
SSES_tpscoresUS <- accuracy(SSES_seas_totalperUS$forecast,ts_US_test[,"Total.Per"])
SS_tpscoresUS <- accuracy(SS_for_totalperUS$mean,ts_US_test[,"Total.Per"])

# Compare the matrix 
tpscoresUS <- as.data.frame(rbind(SANIVE_tpscoresUS, SMA_tpscoresUS, 
                                SES_tpscoresUS, SARIMA_tpscoresUS, ARIMA_tpscoresUS, 
                                ETS_tpscoresUS, ARIMA_Four_tpscoresUS, TBATS_tpscoresUS, 
                                NN_tpscoresUS, SSES_tpscoresUS, SS_tpscoresUS)) 

row.names(tpscoresUS) <- c("SNAIVE", "SMA", "SES", "SARIMA", "ARIMA",
                       "ETS", "ARIMA_FOURIER", "TBATS", "NNETAR",
                       "SSES", "BSM")

tpscoresUS <- tpscoresUS %>%
  mutate(Average = rowMeans(select(., RMSE, MAPE), na.rm = TRUE))
  
# Choose model with lowest error
best_model_index_tpUS <- which.min(tpscoresUS[,"Average"])
cat("The best model by Average is:", row.names(tpscoresUS[best_model_index_tpUS,]))  

# Create Tables 
kbl(tpscoresUS, 
      caption = "Forecast Accuracy for Unemployment Rate (%) Data",
      digits = array(5,ncol(tpscoresUS))) %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  #highlight model with lowest RMSE
  kable_styling(latex_options="striped", stripe_index = which.min(seas_scores[,"Average"]))

# Plot everything together
autoplot(ts_US_test[,"Total.Per"]) +
  autolayer(SNAIVE_deseas_totalperUS, PI=FALSE, series="SNAIVE") + 
  autolayer(SES_deseas_totalperUS, PI=FALSE, series="SES") +
  autolayer(SARIMA_forecast_totalperUS, PI=FALSE, series="SARIMA") +
  autolayer(ARIMA_forecast_totalperUS, PI=FALSE, series="ARIMA") +
  autolayer(ETS_totalperUS, PI=FALSE, series="ETS") +
  autolayer(ARIMA_Four_for_totalperUS, PI=FALSE, series="ARIMA_FOURIER") +
  autolayer(TBATS_for_totalperUS, PI=FALSE, series="TBATS") +
  autolayer(NN_for_totalperUS, PI=FALSE, size=0.7, series="NNETAR") +
  autolayer(SS_for_totalperUS, PI=FALSE, series="BSM") +
  guides(colour=guide_legend(title="Forecast")) # SMA and SSES could not run
```

### Step 5: Forecast for 2025 with the best three models for US (Original)
```{r forecast 2025 US original, warning=FALSE, message=FALSE}
# Set the forecasting period
n_fullUS = 12

# Create the time series to retain full data set
ts_US_fulltrain <- ts(US[,6],
              start=c(year(US$Month[1]), month(US$Month[1])),
              frequency = 12)

# Fit SS with BSM Model 
SS_seas_totalper_fulltrainUS <- StructTS(ts_US_fulltrain,
                    type="BSM",fixed=c(0.01,0.001,0.1,NA)) 

SS_for_totalper_fulltrainUS <- forecast(SS_seas_totalper_fulltrainUS,h=n_fullUS)

# Plot model + observed data
autoplot(ts_US_fulltrain) +
  autolayer(SS_for_totalper_fulltrainUS, series="SS with BSM Model",PI=FALSE)+
  ylab("Forecasted Unemployment Rate (%) in US") 

# Simple Exponential Smoothing Model
decom_totalper_fulltrainUS <- decompose(ts_US_fulltrain)
deseas_totalper_fulltrainUS <- seasadj(decom_totalper_fulltrainUS)
SES_deseas_totalper_fulltrainUS = ses( y = deseas_totalper_fulltrainUS,
                                       h=n_fullUS,    holdout = FALSE,
                                       silent = FALSE)  

# Plot model + observed data
autoplot(ts_US_fulltrain) +
  autolayer(SES_deseas_totalper_fulltrainUS, series="SES Model",PI=FALSE)+
  ylab("Forecasted Unemployment Rate (%) in US") 

# Fit ARIMA 
decom_totalper_fulltrainUS <- decompose(ts_US_fulltrain)
deseas_totalper_fulltrainUS <- seasadj(decom_totalper_fulltrainUS)

ARIMA_totalper_fulltrainUS <- auto.arima(deseas_totalper_fulltrainUS, max.D = 0, 
                               max.P = 0, max.Q = 0)

ARIMA_forecast_totalperUS <- forecast(object = ARIMA_totalper_fulltrainUS, h=n_fullUS)

# Plot model + observed data
autoplot(ts_US_fulltrain) +
  autolayer(ARIMA_forecast_totalperUS, series= "ARIMA Model",PI=FALSE)+
  ylab("Forecasted Unemployment Rate (%) in US") 


# Plot 3 models together 
autoplot(ts_US_fulltrain) +
  autolayer(SS_for_totalper_fulltrainUS, series="SS with BSM Model",PI=FALSE)+
  autolayer(SES_deseas_totalper_fulltrainUS, series="SES Model",PI=FALSE)+
  autolayer(ARIMA_forecast_totalperUS, series= "ARIMA Model",PI=FALSE)+
  ylab("Unemployment Rate (%)") + 
  ggtitle("Forecasted Unemployment Rate (%) in US")
```

## Time Series Analysis for US (Outliers-removed Series)
### Step 1: Transform into time series and set training and testing windows for US (Outliers)
```{r time series US}
# Remove outliers 
ts_USout <- tsclean(ts_US[,"Total.Per"]) 

autoplot(ts_USout, series="Outliers-removed Series") +
  autolayer(ts_US[,"Total.Per"], series="Original Series") +
  ylab("Unemployment Rate (%)") 

# Create a subset for training purpose 
ts_US_trainout <- ts(ts_USout[1:(nobsUS-n_forUS)],
                    start=c(year(US$Month[1]), month(US$Month[1])),
                    frequency = 12)

# Create a subset for testing purpose
ts_US_testout <- ts(ts_USout[(nobsUS - n_forUS + 1):nobsUS],
                   start=c(year(US$Month[start_rowUS]),
                           month(US$Month[start_rowUS])), frequency = 12)

# Plots 
trainUSout <- autoplot(ts_US_trainout) + ylab("Unemployment Rate (%)") +
  ggtitle("Training Window")
testUSout <- autoplot(ts_US_testout) + ylab("Unemployment Rate (%)") +
  ggtitle("Testing Window")
grid.arrange(trainUSout, testUSout, ncol = 2)

par(mfrow=c(1,2))
Acf(ts_US_trainout, lag=40, plot = TRUE, main = "")
Pacf(ts_US_trainout, lag=40, plot = TRUE, main = "")
par(mfrow=c(1,1))
```

### Step 2: Decompose the time series for US (Outliers)
```{r total.per decompose, warning=FALSE, message=FALSE}
# Decompose 
decom_totalper_trainUSout <- decompose(ts_US_trainout)
plot(decom_totalper_trainUSout)

# Deseason 
deseas_totalper_trainUSout <- seasadj(decom_totalper_trainUSout)  
plot(deseas_totalper_trainUSout)

# Run the tests on deseasoned series
print(adf.test(deseas_totalper_trainUSout, alternative = "stationary")) # It is unit root. 
summary(MannKendall(deseas_totalper_trainUSout)) # It has a decreasing trend.

# Run the tests on original series 
print(adf.test(ts_US_trainout, alternative = "stationary")) # It is unit out. 
summary(SeasonalMannKendall(ts_US_trainout)) 
summary(smk.test(ts_US_trainout)) # It has seasonality. 

# Check for any differencing needed 
print(ndiffs(ts_US_trainout))
print(ndiffs(deseas_totalper_trainUSout))
```

### Step 3: Test Time Series Models for US (Outliers)
```{r time series models US, warning=FALSE, message=FALSE}
# Seasonal Naive Model 
SNAIVE_deseas_totalperUSout <- snaive(ts_US_trainout, h=n_forUS)
autoplot(SNAIVE_deseas_totalperUSout)
checkresiduals(SNAIVE_deseas_totalperUSout)

# Simple Moving Average Model
SMA_deseas_totalperUSout <- smooth::sma(y = deseas_totalper_trainUSout, h=n_forUS, 
                                     holdout = FALSE, silent = FALSE) 
summary(SMA_deseas_totalperUSout)
checkresiduals(SMA_deseas_totalperUSout)

# Simple Exponential Smoothing Model
SES_deseas_totalperUSout = ses( y = deseas_totalper_trainUSout, h=n_forUS, 
                             holdout = FALSE, silent = FALSE)  
summary(SES_deseas_totalperUSout)
autoplot(SES_deseas_totalperUSout)
checkresiduals(SES_deseas_totalperUSout)

# SARIMA Model
SARIMA_totalperUSout <- auto.arima(ts_US_trainout)
print(SARIMA_totalperUSout)

SARIMA_forecast_totalperUSout <- forecast(object = SARIMA_totalperUSout, h=n_forUS)
autoplot(SARIMA_forecast_totalperUSout)
checkresiduals(SARIMA_forecast_totalperUSout) # Residuals are not iid.

# Deaseasoned ARIMA Model
ARIMA_totalperUSout <- auto.arima(deseas_totalper_trainUSout, max.D = 0, 
                               max.P = 0, max.Q = 0)
print(ARIMA_totalperUSout)

ARIMA_forecast_totalperUSout <- forecast(object = ARIMA_totalperUSout, h=n_forUS)
autoplot(ARIMA_forecast_totalperUSout)
checkresiduals(ARIMA_forecast_totalperUSout) # Residuals are iid.
 
# STL + ETS Model
ETS_totalperUSout <-  stlf(ts_US_trainout,h=n_forUS)
autoplot(ETS_totalperUSout) 
checkresiduals(ETS_totalperUSout) # Residuals are not iid. 

# ARIMA + FOURIER Model
ARIMA_Four_fit_totalperUSout <- auto.arima(ts_US_trainout, 
                             seasonal=FALSE, lambda=0,
                             xreg=fourier(ts_US_trainout, 
                                          K=3))

ARIMA_Four_for_totalperUSout <- forecast(ARIMA_Four_fit_totalperUSout,
                           xreg=fourier(ts_US_trainout,
                                        K=3, h=n_forUS),
                           h=n_forUS) 

autoplot(ARIMA_Four_for_totalperUSout)
checkresiduals(ARIMA_Four_for_totalperUSout) # Residuals are not iid. 

# TBATS Model 
TBATS_fit_totalperUSout <- tbats(ts_US_trainout)
TBATS_for_totalperUSout <- forecast(TBATS_fit_totalperUSout, h = n_forUS)
autoplot(TBATS_for_totalperUSout) 
checkresiduals(TBATS_fit_totalperUSout) # Residuals are not iid. 

# Neural Network Model 
NN_fit_totalperUSout <- nnetar(ts_US_trainout,
                 p=3, P=0,
                 xreg=fourier(ts_US_trainout, K=3))

NN_for_totalperUSout <- forecast(NN_fit_totalperUSout, 
                   h=n_forUS,
                   xreg=fourier(ts_US_trainout, 
                                          K=3,h=n_forUS))

autoplot(NN_for_totalperUSout)
checkresiduals(NN_fit_totalperUSout) # Residuals are not iid. 

## State Space Exponential Smoothing Model
SSES_seas_totalperUSout <- es(ts_US_trainout,
                         model="ZZZ", h=n_forUS, holdout=FALSE)
checkresiduals(SSES_seas_totalperUSout) # Residuals are not iid.

## State Space with BSM Model
SS_seas_totalperUSout <- StructTS(ts_US_trainout,
                    type="BSM",fixed=c(0.01,0.001,0.1,NA)) 

SS_for_totalperUSout <- forecast(SS_seas_totalperUSout,h=n_forUS)

plot(SS_for_totalperUSout)
checkresiduals(SS_seas_totalperUSout) # Residuals are not iid. 
```

### Step 4: Performance check for US (Outliers)
```{r accuracy, warning=FALSE, message=FALSE}
# Check accuracy of the models
SANIVE_tpscoresUSout <- accuracy(SNAIVE_deseas_totalperUSout$mean,ts_US_testout)  
SMA_tpscoresUSout <- accuracy(SMA_deseas_totalperUSout$forecast,ts_US_testout)  
SES_tpscoresUSout <- accuracy(SES_deseas_totalperUSout$mean,ts_US_testout)
SARIMA_tpscoresUSout <- accuracy(SARIMA_forecast_totalperUSout$mean,ts_US_testout)
ARIMA_tpscoresUSout <- accuracy(ARIMA_forecast_totalperUSout$mean,ts_US_testout)
ETS_tpscoresUSout <- accuracy(ETS_totalperUSout$mean,ts_US_testout)
ARIMA_Four_tpscoresUSout <- accuracy(ARIMA_Four_for_totalperUSout$mean,ts_US_testout)
TBATS_tpscoresUSout <- accuracy(TBATS_for_totalperUSout$mean,ts_US_testout)
NN_tpscoresUSout <- accuracy(NN_for_totalperUSout$mean,ts_US_testout)
SSES_tpscoresUSout <- accuracy(SSES_seas_totalperUSout$forecast,ts_US_testout)
SS_tpscoresUSout <- accuracy(SS_for_totalperUSout$mean,ts_US_testout)

# Compare the matrix 
tpscoresUSout <- as.data.frame(rbind(SANIVE_tpscoresUSout, SMA_tpscoresUSout, 
                                SES_tpscoresUSout, SARIMA_tpscoresUSout, ARIMA_tpscoresUSout, 
                                ETS_tpscoresUSout, ARIMA_Four_tpscoresUSout, TBATS_tpscoresUSout, 
                                NN_tpscoresUSout, SSES_tpscoresUSout, SS_tpscoresUSout)) 

row.names(tpscoresUSout) <- c("SNAIVE", "SMA", "SES", "SARIMA", "ARIMA",
                       "ETS", "ARIMA_FOURIER", "TBATS", "NNETAR",
                       "SSES", "BSM")

tpscoresUSout <- tpscoresUSout %>%
  mutate(Average = rowMeans(select(., RMSE, MAPE), na.rm = TRUE))

# Choose model with lowest error
best_model_index_tpUSout <- which.min(tpscoresUSout[,"Average"])
cat("The best model by Average is:", row.names(tpscoresUSout[best_model_index_tpUSout,]))  

# Create Tables 
kbl(tpscoresUSout, 
      caption = "Forecast Accuracy for Unemployment Rate (%) Data",
      digits = array(5,ncol(tpscoresUSout))) %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  #highlight model with lowest RMSE
  kable_styling(latex_options="striped", stripe_index = which.min(seas_scores[,"Average"]))

# Plot everything together
autoplot(ts_US_testout) +
  autolayer(SNAIVE_deseas_totalperUSout, PI=FALSE, series="SNAIVE") + 
  autolayer(SES_deseas_totalperUSout, PI=FALSE, series="SES") +
  autolayer(SARIMA_forecast_totalperUSout, PI=FALSE, series="SARIMA") +
  autolayer(ARIMA_forecast_totalperUSout, PI=FALSE, series="ARIMA") +
  autolayer(ETS_totalperUSout, PI=FALSE, series="ETS") +
  autolayer(ARIMA_Four_for_totalperUSout, PI=FALSE, series="ARIMA_FOURIER") +
  autolayer(TBATS_for_totalperUSout, PI=FALSE, series="TBATS") +
  autolayer(NN_for_totalperUSout, PI=FALSE, series="NNETAR") +
  autolayer(SS_for_totalperUSout, PI=FALSE, series="BSM") +
  guides(colour=guide_legend(title="Forecast")) # SMA and SSES could not run
```

### Step 5: Forecast for 2025 with the best three models for US (Outliers)
```{r forecast 2025 US, warning=FALSE, message=FALSE}
# Set the forecasting period
n_fullUS = 12

# Create the time series to retain full data set
ts_US_fulltrainout <- ts(ts_USout,
              start=c(year(US$Month[1]), month(US$Month[1])),
              frequency = 12)

# Fit SS with BSM Model 
SS_seas_totalper_fulltrainUSout <- StructTS(ts_US_fulltrainout,
                    type="BSM",fixed=c(0.01,0.001,0.1,NA)) 

SS_for_totalper_fulltrainUSout <- forecast(SS_seas_totalper_fulltrainUSout,h=n_fullUS)

# Plot model + observed data
autoplot(ts_US_fulltrainout) +
  autolayer(SS_for_totalper_fulltrainUSout, series="SS with BSM Model",PI=FALSE)+
  ylab("Forecasted Unemployment Rate (%) in US") 

# Fit Neural Network Model 
NN_fit_totalper_fulltrainUSout <- nnetar(ts_US_fulltrainout,
                 p=3, P=0,
                 xreg=fourier(ts_US_fulltrainout, K=3))

NN_for_totalper_fulltrainUSout <- forecast(NN_fit_totalper_fulltrainUSout, 
                   h=n_fullUS,
                   xreg=fourier(ts_US_fulltrainout, 
                                          K=3,h=n_fullUS))

# Plot model + observed data
autoplot(ts_US_fulltrainout) +
  autolayer(NN_for_totalper_fulltrainUSout, series="NNETAR",PI=FALSE)+
  ylab("Forecasted Unemployment Rate (%) in US")

# Fit STL + ETS Model
ETS_totalper_fulltrainUSout <-  stlf(ts_US_fulltrainout,h=n_fullUS)

# Plot model + observed data
autoplot(ts_US_fulltrainout) +
  autolayer(ETS_totalper_fulltrainUSout, series="ETS",PI=FALSE)+
  ylab("Forecasted Unemployment Rate (%) in US")

# Plot 4 models together 
autoplot(ts_US_fulltrain) +
  autolayer(SS_for_totalper_fulltrainUSout, series="SS with BSM Model",PI=FALSE)+
  autolayer(NN_for_totalper_fulltrainUSout, series="NNETAR",PI=FALSE)+
  autolayer(ETS_totalper_fulltrainUSout, series="ETS",PI=FALSE)+
  ylab("Unemployment Rate (%)") + 
  ggtitle("Forecasted Unemployment Rate (%) in US (Outliers Removed)")
```

### Step 6: The average of 3 forecasts
```{r}

# --- 1. Calculate the average forecast of the three models ---
# Extract predicted means
bsm_fc <- as.numeric(SS_for_totalper_fulltrainUSout$mean)
nnar_fc <- as.numeric(NN_for_totalper_fulltrainUSout$mean)
ets_fc <- as.numeric(ETS_totalper_fulltrainUSout$mean)

# Compute average forecast
avg_fc <- (bsm_fc + nnar_fc + ets_fc) / 3

# Create date index for forecast
start_date <- as.Date(paste0(end(US$Month)[1] + 1, "-", end(US$Month)[2], "-01"))  # 1 month after last
forecast_dates <- seq(from = as.Date("2025-01-01"), by = "month", length.out = n_fullUS)

# Create forecast dataframe
forecast_df <- tibble(
  Month = forecast_dates,
  BSM = bsm_fc,
  NNAR = nnar_fc,
  ETS = ets_fc,
  Avg_Forecast = avg_fc
)

# --- 2. Export forecast to Excel ---
write.xlsx(forecast_df, "./Output/Forecast Average/Forecast_Average_US.xlsx")

# --- 3. Plot actuals (from 2020) and forecast average ---
# Extract actual values after 2020
us_actual_2020on <- US %>%
  filter(Month >= as.Date("2020-01-01")) %>%
  select(Month, Actual = Total.Per)


# Combine with forecast
plot_df <- bind_rows(
  tibble(Month = us_actual_2020on$Month,
         Value = us_actual_2020on$Actual,
         Type = "Actual"),
  tibble(Month = forecast_dates,
         Value = avg_fc,
         Type = "Avg Forecast")
)

# --- 4. Plot actuals (from 2022) and forecast average ---
# Extract actual values after 2022
us_actual_2022on <- US %>%
  filter(Month >= as.Date("2022-01-01")) %>%
  select(Month, Actual = Total.Per)


# Combine with forecast
plot_df2022 <- bind_rows(
  tibble(Month = us_actual_2022on$Month,
         Value = us_actual_2022on$Actual,
         Type = "Actual"),
  tibble(Month = forecast_dates,
         Value = avg_fc,
         Type = "Avg Forecast")
)


# Plot
ggplot(plot_df2022, aes(x = Month, y = Value, color = Type, group = 1)) +
  geom_line(size = 1) +
  scale_color_manual(values = c("Actual" = "black", "Avg Forecast" = "steelblue")) +
  labs(
    title = "US Unemployment Rate: Actual (2018~) + Avg Forecast (2025)",
    y = "Unemployment Rate (%)",
    x = "Month",
    color = "Legend"
  ) +
  theme_minimal() + theme(legend.position = "bottom")
```

