---
title: "Data Wrangling"
author: "Aye Nyein Thu"
date: "2025-03-24"
output: html_document
editor_options: 
  chunk_output_type: console
---

## Setting R code chunk options
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

## Loading packages and initializing
```{r packages, warning=FALSE, message=FALSE}
# Load required packages
library(readxl)
library(openxlsx)
library(writexl)
library(dplyr)
library(lubridate)
library(ggplot2)
library(cowplot)
library(forecast)  
library(Kendall)
library(tseries)
library(outliers)
library(tidyverse)
library(smooth)
library(trend)
library(kableExtra)
library(tidyr)
library(gt) #install.packages("gt")
library(gridExtra) #install.packages("gridExtra")
library(zoo)

# Check working directory
getwd()
```

## Importing and Wrangling Data
```{r importing data, results='hide', warning=FALSE, message=FALSE}
## Raw Data Set: Unemployment Rate by Age (Thousands)
# Import data set
UEAge.Thou <- read_excel(
  path="./Data/Raw/UE_Age(Thousands).xlsx", sheet = "Sheet1", col_names = TRUE)

# Format data set
UEAge.Thou_Processed <- UEAge.Thou %>%
  mutate(
    Month = ym(sub("M", "-", Month)), 
    Age15to24.Thou = as.numeric(`15-24`), 
    Age25above.Thou = as.numeric(`25+`),   
    AgeTotal.Thou = as.numeric(`15+`)) %>% 
  rename(Country="Reference area") %>% 
  select(Country,Month,Age15to24.Thou, Age25above.Thou, AgeTotal.Thou) %>% 
  arrange(Country, Month)

## Raw Data Set: Unemployment Rate by Age (%)
# Import data set
UEAge.Per <- read_excel(
  path="./Data/Raw/UE_Age(%).xlsx", sheet = "Sheet1", col_names = TRUE)

# Format data set
UEAge.Per_Processed <- UEAge.Per %>%
  mutate(
    Month = ym(sub("M", "-", Month)), 
    Age15to24.Per = as.numeric(`15-24`), 
    Age25above.Per = as.numeric(`25+`),   
    AgeTotal.Per = as.numeric(`15+`)) %>% 
  rename(Country="Reference area") %>% 
  select(Country,Month,Age15to24.Per, Age25above.Per, AgeTotal.Per) %>% 
  arrange(Country, Month)

## Raw Data Set: Unemployment Rate by Gender (Thousands)
# Import data set
UEGender.Thou <- read_excel(
  path="./Data/Raw/UE_Gender(Thousands).xlsx", sheet = "Sheet1", col_names = TRUE)

# Format data set
UEGender.Thou_Processed <- UEGender.Thou %>%
  mutate(
    Month = ym(sub("M", "-", Month)), 
    Female.Thou = as.numeric(Female), 
    Male.Thou = as.numeric(Male),   
    Total.Thou = as.numeric(Total)) %>% 
  rename(Country="Reference area") %>% 
  select(Country,Month,Female.Thou, Male.Thou, Total.Thou) %>% 
  arrange(Country, Month)

## Raw Data Set: Unemployment Rate by Gender (%)
# Import data set
UEGender.Per <- read_excel(
  path="./Data/Raw/UE_Gender(%).xlsx", sheet = "Sheet1", col_names = TRUE)

# Format data set
UEGender.Per_Processed <- UEGender.Per %>%
  mutate(
    Month = ym(sub("M", "-", Month)), 
    Female.Per = as.numeric(Female), 
    Male.Per = as.numeric(Male),   
    Total.Per = as.numeric(Total)) %>% 
  rename(Country="Reference area") %>% 
  select(Country,Month,Female.Per, Male.Per, Total.Per) %>% 
  arrange(Country, Month)
```

```{r data wrangling Peru, warning=FALSE, message=FALSE}
# Combine all processed data sets 
UE_Countries <- UEAge.Thou_Processed %>% 
  left_join(UEAge.Per_Processed, by=c("Country", "Month")) %>% 
  left_join(UEGender.Thou_Processed, by=c("Country", "Month")) %>% 
  left_join(UEGender.Per_Processed, by=c("Country", "Month")) 

# Extract Peru Data 
Peru <- UE_Countries %>% 
  filter(Country == "Peru") %>% 
  select(-Country, AgeTotal.Per, AgeTotal.Thou) %>% 
  select(Month, Age15to24.Per, Age25above.Per, Female.Per, Male.Per,
         Total.Per, Age15to24.Thou, Age25above.Thou, Female.Thou, Male.Thou,
         Total.Thou) 

# Check Missing Value 
sum(is.na(Peru))
```

```{r data wrangling global, warning=FALSE, message=FALSE}
# Generate global unemployment data using simple average 
UE_Global <- UE_Countries %>%
  group_by(Month) %>%
  summarise(
    Age15to24.Thou = mean(`Age15to24.Thou`, na.rm = TRUE),
    Age25above.Thou = mean(`Age25above.Thou`, na.rm = TRUE), 
    AgeTotal.Thou = mean(`AgeTotal.Thou`, na.rm = TRUE), 
    Age15to24.Per = mean(`Age15to24.Thou`, na.rm = TRUE),
    Age25above.Per = mean(`Age25above.Per`, na.rm = TRUE), 
    AgeTotal.Per = mean(`AgeTotal.Per`, na.rm = TRUE), 
    Female.Thou = mean(`Female.Thou`, na.rm = TRUE),
    Male.Thou = mean(`Male.Thou`, na.rm = TRUE),
    Total.Thou = mean(`Total.Thou`, na.rm = TRUE),
    Female.Per = mean(`Female.Per`, na.rm = TRUE),
    Male.Per = mean(`Male.Per`, na.rm = TRUE),
    Total.Per = mean(`Total.Per`, na.rm = TRUE))

# Generate global unemployment data using weighted average 
UE_Global_Weighted <- UE_Countries %>%
  filter(apply(UE_Countries[, 3:14], 1, function(x) all(!is.na(x)))) %>%  
  group_by(Month) %>%
  summarise(
    Age15to24.Per = sum(Age15to24.Thou) / sum(Age15to24.Thou / Age15to24.Per), 
    Age25above.Per = sum(Age25above.Thou) / sum(Age25above.Thou / Age25above.Per),
    Female.Per = sum(Female.Thou) / sum(Female.Thou / Female.Per),
    Male.Per = sum(Male.Thou) / sum(Male.Thou / Male.Per),
    Total.Per = sum(Total.Thou) / sum(Total.Thou / Total.Per),
    
    Age15to24.Thou = mean(Age15to24.Thou, na.rm = TRUE),
    Age25above.Thou = mean(Age25above.Thou, na.rm = TRUE),
    Female.Thou = mean(Female.Thou, na.rm = TRUE),
    Male.Thou = mean(Male.Thou, na.rm = TRUE),
    Total.Thou = mean(Total.Thou, na.rm = TRUE)
  )

# Generate weighted global unemployment yearly
UE_Global_Yearly <- UE_Global_Weighted %>% 
  mutate(Year = year(Month)) %>%
  group_by(Year) %>% 
  summarize(across(where(is.numeric), mean)) #%>% 
  #filter(Year >= 2000 & Year <= 2024)

```

## Descriptive Statistics 
### Global Unemployment Status
```{r global, warning=FALSE, message=FALSE}
# Check total no. of countries 
print(length(unique(UE_Countries$Country)))

# Plot global unemployment rate for simple average vs. weighted average 
UE_Global_Combined <- bind_rows(
  UE_Global %>% mutate(Source = "Simple Average"),
  UE_Global_Weighted %>% mutate(Source = "Weighted Average")
)

ggplot(UE_Global_Combined, aes(x = Month, y = Total.Per, color = Source)) +
  geom_line() +               
  labs(title = "Global Unemployment Rate: Simple Vs. Weighted Average",
       x = "Month", y = "Global Unemployment Rate (%)", color = "Source") 

# Plots
# Global Unemployment Rate 
ggplot(UE_Global_Yearly, aes(x = Year, y = Total.Per)) +
  geom_line() +               
  labs(title = "Global Unemployment Rate",
       x = "Year", y = "Global Unemployment Rate (%)")

# Global Unemployment Rate by Age
ggplot(UE_Global_Yearly, aes(x = Year)) +
  geom_line(aes(y = Age15to24.Per, color = "Age 15-24")) +
  geom_line(aes(y = Age25above.Per, color = "Age 25+")) +
  labs(title = "Global Unemployment Rate by Age",
       x = "Year", y = "Global Unemployment Rate (%)", color = "Age Group") +
  scale_color_manual(values = c("Age 15-24" = "blue", "Age 25+" = "red"))

# Global Unemployment Rate by Sex
ggplot(UE_Global_Yearly, aes(x = Year)) +
  geom_line(aes(y = Female.Per, color = "Female")) +
  geom_line(aes(y = Male.Per, color = "Male")) +
  labs(title = "Global Unemployment Rate by Sex",
       x = "Year", y = "Global Unemployment Rate (%)", color = "Sex Group") +
  scale_color_manual(values = c("Female" = "blue", "Male" = "red"))
```

### Peru Unemployment Status
```{r peru, warning=FALSE, message=FALSE}
# First plot: Unemployment Rate (Thousands)
p1 <- ggplot(Peru, aes(x = Month, y = Total.Thou)) +
  geom_line(color = "blue", alpha = 1) +
  labs(title = "Unemployment in Thousands", x = "Year", y = "Thousands") 

# Second plot: Unemployment Rate (%)
p2 <- ggplot(Peru, aes(x = Month, y = Total.Per)) +
  geom_line(color = "red", alpha = 1) +
  labs(title = "Unemployment Rate (%)", x = "Year", y = "Percentage") 

# Arrange plots side by side
grid.arrange(p1, p2, ncol = 2)

# Generate Laborforce Participation 
Peru_LF <- Peru %>% 
  mutate(LF.Part = (Total.Thou / Total.Per) * 100) 

p3 <- ggplot(Peru_LF, aes(x = Month, y = LF.Part)) +
  geom_line(color = "orange", alpha = 1) +
  labs(title = "Laborforce Participation Rate (%)", x = "Year", y = "Thousands")

# Arrange plots side by side
grid.arrange(p1, p2, p3, ncol = 3)

# Take a few countries out for comparison
UE_Comparison <- UE_Countries %>% 
  filter(Country %in% c("Peru", "Colombia", "Mexico", "Chile")) %>% 
  mutate(Month = as.Date(Month)) 

# Plot total unemployment rate for each countries
ggplot(UE_Comparison, aes(x = Month, y = Total.Per, color = Country)) +
  geom_line(size = 1.2) +  
  geom_point(size = 1.5, alpha = 0.7) + 
  theme_gray(base_size = 14) +  
  labs(title = "Unemployment Rate by Countries (2001-2024)",
    x = "Year", y = "Unemployment Rate (%)", color = "Country") +
  theme(plot.title = element_text(face = "bold", size = 16),
    axis.title.y = element_text(size = 14), axis.text = element_text(size = 12),
    legend.position = "right")

# Peru Unemployment Rate by Age Group 
ggplot(Peru, aes(x = Month)) +
  geom_line(aes(y = Age15to24.Per, color = "Age 15-24"), size = 1.2) +
  geom_line(aes(y = Age25above.Per, color = "Age 25+"), size = 1.2) +
  geom_point(aes(y = Age15to24.Per, color = "Age 15-24"), size = 2, alpha = 0.7) +
  geom_point(aes(y = Age25above.Per, color = "Age 25+"), size = 2, alpha = 0.7) +
  labs(title = "Peru Unemployment Rate by Age",
    x = "Year", y = "Unemployment Rate (%)", color = "Age Group") +
  scale_color_manual(values = c("Age 15-24" = "blue", "Age 25+" = "red")) +
  theme_gray(base_size = 14) +  
  theme(plot.title = element_text(face = "bold", size = 16),
    axis.title.x = element_text(size = 14), axis.title.y = element_text(size = 14),
    axis.text = element_text(size = 12), legend.position = "bottom") 

# Peru Unemployment Rate by Sex Group 
ggplot(Peru, aes(x = Month)) +
  geom_line(aes(y = Female.Per, color = "Female"), size = 1.2) +
  geom_line(aes(y = Male.Per, color = "Male"), size = 1.2) +
  geom_point(aes(y = Female.Per, color = "Female"), size = 2, alpha = 0.7) +
  geom_point(aes(y = Male.Per, color = "Male"), size = 2, alpha = 0.7) +
  labs(title = "Peru Unemployment Rate by Sex",
    x = "Year", y = "Unemployment Rate (%)", color = "Sex Group") +
  scale_color_manual(values = c("Female" = "blue", "Male" = "red")) +
  theme_gray(base_size = 14) +  # Minimal theme for cleaner look
  theme(plot.title = element_text(face = "bold", size = 16),
    axis.title.x = element_text(size = 14), axis.title.y = element_text(size = 14),
    axis.text = element_text(size = 12), legend.position = "bottom")
```

```{r missing month handling}
# Check regularity of time index
diff_date <- diff(Peru$Month)
table(diff_date)  # There's one missing month.

# Create full monthly sequence
full_month_seq <- data.frame(Month = seq.Date(from = min(Peru$Month),
                                              to = max(Peru$Month),
                                              by = "month"))

# Join with original to find missing months
missing_months <- full_month_seq %>%
  anti_join(Peru, by = "Month")

print(missing_months)  # Missing month is 2012-09-01

# Fill in missing months with NA and interpolate linearly
Peru <- full_month_seq %>%
  left_join(Peru, by = "Month") %>% 
  mutate(across(where(is.numeric), ~na.approx(.x, na.rm = FALSE)))
```

```{r summary stats table}
# Generate Summary Statistics
summary_table <- Peru %>%
  select(-Month) %>%  # Exclude Month column
  summarise(across(where(is.numeric), 
                   list(Mean = ~ mean(.x),
                        SD = ~ sd(.x),
                        Min = ~ min(.x),
                        Max = ~ max(.x),
                        N = ~ sum(!is.na(.x))))) %>%
  pivot_longer(everything(), names_to = c("Variable", ".value"), names_sep = "_") %>% 
  gt() %>%
  tab_header(title = "Summary Statistics of Unemployment Data in Peru",
    subtitle = "Monthly Data (2001-2024)") %>%
  fmt_number(columns = 2:6, decimals = 2) %>%
  cols_label(Variable = "Indicator", Mean = "Mean", SD = "Standard Deviation",
    Min = "Min", Max = "Max", N = "Observations") %>%
  tab_options(table.font.size = px(14),
    heading.title.font.size = px(18), heading.subtitle.font.size = px(14))

print(summary_table)

# Check outliers 
outlier(Peru) 
grubbs.test(Peru$Age15to24.Thou) # This is an outlier. 
grubbs.test(Peru$Age25above.Thou) # This is an outlier. 
grubbs.test(Peru$Age15to24.Per) 
grubbs.test(Peru$Age25above.Per)
grubbs.test(Peru$Female.Thou) # This is an outlier. 
grubbs.test(Peru$Male.Thou) # This is an outlier. 
grubbs.test(Peru$Female.Per)
grubbs.test(Peru$Male.Per)
grubbs.test(Peru$Total.Thou) # This is an outlier. 
grubbs.test(Peru$Total.Per)

# Check the box plot for total unemployment 
boxplot(Peru$Total.Per,
        main = "Boxplot: Peru Unemployment Rate (%)",
        horizontal = TRUE, 
        col = "lightblue")

# Find the row where Total.Per is the outlier (10.8%)
outlier_row <- Peru %>%
  filter(Total.Per == max(Total.Per, na.rm = TRUE))

print(outlier_row) # Highest Value is 2005-02-01. 
```

## Transforming into time series and Setting training and testing windows
```{r time series}
# Transform into time series
ts_Peru <- ts(Peru[,2:11],
              start=c(year(Peru$Month[1]), month(Peru$Month[1])),
              frequency = 12)

# Set the period
nobs = nrow(Peru)
n_for = 12

# Create a subset for training purpose 
ts_Peru_train <- ts(Peru[1:(nobs-n_for),2:11],
                    start=c(year(Peru$Month[1]), month(Peru$Month[1])),
                    frequency = 12)

# Create a subset for testing purpose
start_row = nobs - n_for + 1
ts_Peru_test <- ts(Peru[(nobs-n_for+1):nobs,2:11],
                   start=c(year(Peru$Month[start_row]),
                           month(Peru$Month[start_row])), frequency = 12)

# Plots 
train <- autoplot(ts_Peru_train[,"Total.Per"]) + ylab("Unemployment Rate (%)") +
  ggtitle("Training Window")
test <- autoplot(ts_Peru_test[,"Total.Per"]) + ylab("Unemployment Rate (%)") +
  ggtitle("Testing Window")
grid.arrange(train, test, ncol = 2)

par(mfrow=c(1,2))
Acf(ts_Peru_train[,"Total.Per"], lag=40, plot = TRUE, main = "")
Pacf(ts_Peru_train[,"Total.Per"], lag=40, plot = TRUE, main = "")
par(mfrow=c(1,1))
```

## Decompose the time series 
```{r total.per decompose, warning=FALSE, message=FALSE}
# Decompose 
decom_totalper_train <- decompose(ts_Peru_train[,"Total.Per"])
plot(decom_totalper_train)

# Deseason 
deseas_totalper_train <- seasadj(decom_totalper_train)  
plot(deseas_totalper_train)

# Run the tests on deseasoned series
print(adf.test(deseas_totalper_train, alternative = "stationary")) # It is unit root. 
summary(MannKendall(deseas_totalper_train)) # It has a decreasing trend.

# Run the tests on original series 
print(adf.test(ts_Peru_train[,"Total.Per"], alternative = "stationary")) # It is stationary. 
summary(SeasonalMannKendall(ts_Peru_train[,"Total.Per"])) 
summary(smk.test(ts_Peru_train[,"Total.Per"])) # It has seasonality. 

# Check for any differencing needed 
print(ndiffs(ts_Peru_train[,"Total.Per"]))
print(ndiffs(deseas_totalper_train))
```

## Testing Time Series Models 
```{r time series models, warning=FALSE, message=FALSE}
# Seasonal Naive Model 
SNAIVE_deseas_totalper <- snaive(ts_Peru_train[,"Total.Per"], h=n_for)
autoplot(SNAIVE_deseas_totalper)
checkresiduals(SNAIVE_deseas_totalper)

# Simple Moving Average Model
SMA_deseas_totalper <- smooth::sma(y = deseas_totalper_train, h=n_for, holdout = FALSE, silent = FALSE) 
summary(SMA_deseas_totalper)
checkresiduals(SMA_deseas_totalper)

# Simple Exponential Smoothing Model
SES_deseas_totalper = ses( y = deseas_totalper_train, h=n_for, holdout = FALSE, silent = FALSE)  
summary(SES_deseas_totalper)
autoplot(SES_deseas_totalper)
checkresiduals(SES_deseas_totalper)

# SARIMA Model
SARIMA_totalper <- auto.arima(ts_Peru_train[,"Total.Per"])
print(SARIMA_totalper)

SARIMA_forecast_totalper <- forecast(object = SARIMA_totalper, h=n_for)
autoplot(SARIMA_forecast_totalper)
checkresiduals(SARIMA_forecast_totalper) # Residuals are iid.

# Deaseasoned ARIMA Model
ARIMA_totalper <- auto.arima(deseas_totalper_train, max.D = 0, max.P = 0, max.Q = 0)
print(ARIMA_totalper)

ARIMA_forecast_totalper <- forecast(object = ARIMA_totalper, h=n_for)
autoplot(ARIMA_forecast_totalper)
checkresiduals(ARIMA_forecast_totalper) # Residuals are iid.
 
# STL + ETS Model
ETS_totalper <-  stlf(ts_Peru_train[,"Total.Per"],h=n_for)
autoplot(ETS_totalper) 
checkresiduals(ETS_totalper) # Residuals are not iid. 

# ARIMA + FOURIER Model
ARIMA_Four_fit_totalper <- auto.arima(ts_Peru_train[,"Total.Per"], 
                             seasonal=FALSE, lambda=0,
                             xreg=fourier(ts_Peru_train[,"Total.Per"], 
                                          K=3))

ARIMA_Four_for_totalper <- forecast(ARIMA_Four_fit_totalper,
                           xreg=fourier(ts_Peru_train[,"Total.Per"],
                                        K=3, h=n_for),
                           h=n_for) 

autoplot(ARIMA_Four_for_totalper)
checkresiduals(ARIMA_Four_for_totalper) # Better fit

# TBATS Model 
TBATS_fit_totalper <- tbats(ts_Peru_train[,"Total.Per"])
TBATS_for_totalper <- forecast(TBATS_fit_totalper, h = n_for)
autoplot(TBATS_for_totalper) 
checkresiduals(TBATS_fit_totalper) # Better fit

# Neural Network Model 
NN_fit_totalper <- nnetar(ts_Peru_train[,"Total.Per"],
                 p=3, P=0,
                 xreg=fourier(ts_Peru_train[,"Total.Per"], K=3))

NN_for_totalper <- forecast(NN_fit_totalper, 
                   h=n_for,
                   xreg=fourier(ts_Peru_train[,"Total.Per"], 
                                          K=3,h=n_for))

autoplot(NN_for_totalper)
checkresiduals(NN_fit_totalper) # Residuals are not iid.

## State Space Exponential Smoothing Model
SSES_seas_totalper <- es(ts_Peru_train[,"Total.Per"],
                         model="ZZZ", h=n_for, holdout=FALSE)
checkresiduals(SSES_seas_totalper) # Residuals are not iid.

## State Space with BSM Model
SS_seas_totalper <- StructTS(ts_Peru_train[,"Total.Per"],
                    type="BSM",fixed=c(0.3,0.01,0.1,NA)) 

SS_for_totalper <- forecast(SS_seas_totalper,h=n_for)

plot(SS_for_totalper)
checkresiduals(SS_seas_totalper) # Residuals are not iid. 
```

## Performance Check
```{r accuracy, warning=FALSE, message=FALSE}
# Check accuracy of the models
SANIVE_tpscores <- accuracy(SNAIVE_deseas_totalper $mean,ts_Peru_test[,"Total.Per"])  
SMA_tpscores <- accuracy(SMA_deseas_totalper$forecast,ts_Peru_test[,"Total.Per"])  
SES_tpscores <- accuracy(SES_deseas_totalper$mean,ts_Peru_test[,"Total.Per"])
SARIMA_tpscores <- accuracy(SARIMA_forecast_totalper$mean,ts_Peru_test[,"Total.Per"])
ARIMA_tpscores <- accuracy(ARIMA_forecast_totalper$mean,ts_Peru_test[,"Total.Per"])
ETS_tpscores <- accuracy(ETS_totalper$mean,ts_Peru_test[,"Total.Per"])
ARIMA_Four_tpscores <- accuracy(ARIMA_Four_for_totalper$mean,ts_Peru_test[,"Total.Per"])
TBATS_tpscores <- accuracy(TBATS_for_totalper$mean,ts_Peru_test[,"Total.Per"])
NN_tpscores <- accuracy(NN_for_totalper$mean,ts_Peru_test[,"Total.Per"])
SSES_tpscores <- accuracy(SSES_seas_totalper$forecast,ts_Peru_test[,"Total.Per"])
SS_tpscores <- accuracy(SS_for_totalper$mean,ts_Peru_test[,"Total.Per"])

# Compare the matrix 
tpscores <- as.data.frame(rbind(SANIVE_tpscores, SMA_tpscores, 
                                SES_tpscores, SARIMA_tpscores, ARIMA_tpscores, 
                                ETS_tpscores, ARIMA_Four_tpscores, TBATS_tpscores, 
                                NN_tpscores, SSES_tpscores, SS_tpscores)) %>%
  mutate(Average = rowMeans(., na.rm = TRUE))

row.names(tpscores) <- c("SNAIVE", "SMA", "SES", "SARIMA", "ARIMA",
                       "ETS", "ARIMA_FOURIER", "TBATS", "NNETAR",
                       "SSES", "BSM")

# Choose model with lowest error
best_model_index_tp <- which.min(tpscores[,"Average"])
cat("The best model by Average is:", row.names(tpscores[best_model_index_tp,]))  

# Create Tables 
kbl(tpscores, 
      caption = "Forecast Accuracy for Unemployment Rate (%) Data",
      digits = array(5,ncol(tpscores))) %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  #highlight model with lowest RMSE
  kable_styling(latex_options="striped", stripe_index = which.min(seas_scores[,"Average"]))

# Plot everything together
autoplot(ts_Peru_test[,"Total.Per"]) +
  autolayer(SNAIVE_deseas_totalper, PI=FALSE, series="SNAIVE") + 
  autolayer(SES_deseas_totalper, PI=FALSE, series="SES") +
  autolayer(SARIMA_forecast_totalper, PI=FALSE, series="SARIMA") +
  autolayer(ARIMA_forecast_totalper, PI=FALSE, series="ARIMA") +
  autolayer(ETS_totalper, PI=FALSE, series="ETS") +
  autolayer(ARIMA_Four_for_totalper, PI=FALSE, series="ARIMA_FOURIER") +
  autolayer(TBATS_for_totalper, PI=FALSE, series="TBATS") +
  autolayer(NN_for_totalper, PI=FALSE, series="NNETAR") +
  autolayer(SS_for_totalper, PI=FALSE, series="BSM") +
  guides(colour=guide_legend(title="Forecast")) # SMA and SSES could not run
```

## Forecasting for 2025 with the best three models 
```{r forecast 2025, warning=FALSE, message=FALSE}
# State Space with BSM Model 
n_full = 52

# Create the time series to retain full data set
ts_Peru_fulltrain <- ts(Peru[,6],
              start=c(year(Peru$Month[1]), month(Peru$Month[1])),
              frequency = 12)

# Fit SSES Model 
SSES_seas_totalper_fulltrain <- es(ts_Peru_fulltrain,
                         model="ZZZ", h=n_full, holdout=FALSE)

SSES_for_totalper_fulltrain <- forecast(SSES_seas_totalper_fulltrain, h=n_full)

# Plot model + observed data
autoplot(ts_Peru_fulltrain) +
  autolayer(SSES_for_totalper_fulltrain, series="SSES",PI=FALSE)+
  ylab("Forecasted Unemployment Rate (%) in Peru") 

# Fit ETS Model 
ETS_seas_totalper_fulltrain <-  stlf(ts_Peru_fulltrain,h=n_full)
ETS_for_totalper_fulltrain <- forecast(ETS_seas_totalper_fulltrain, h=n_full)

# Plot model + observed data
autoplot(ts_Peru_fulltrain) +
  autolayer(ETS_for_totalper_fulltrain, series="ETS",PI=FALSE)+
  ylab("Forecasted Unemployment Rate (%) in Peru")

# Fit SS with BSM Model 
SS_seas_totalper_fulltrain <- StructTS(ts_Peru_fulltrain,
                    type="BSM",fixed=c(0.3,0.01,0.1,NA)) 

SS_for_totalper_fulltrain <- forecast(SS_seas_totalper_fulltrain,h=n_full)

# Plot model + observed data
autoplot(ts_Peru_fulltrain) +
  autolayer(SS_for_totalper_fulltrain, series="SS with BSM Model",PI=FALSE)+
  ylab("Forecasted Unemployment Rate (%) in Peru") 
```

